
---
title: "Detection Probability Sensitivity Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
    df_print: tibble
---

```{r setup, include=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(readr)
library(knitr)
library(kableExtra)
source("R/plot_detection_prob.R")
```

# Overview

This report presents results from a two-part sensitivity analysis using the Generalized Joint Attribute Model (GJAM) to assess plant species abundance patterns across NEON sites. The approach directly addresses core questions from the original analysis (2018–2019), using updated data and a Bayesian joint species distribution model.

# Analytical Goals Revisited

These analyses respond directly to goals 2 and 3 outlined in the original sensitivity analysis report:

## Goal 2: Detecting Interannual Variation
> Can we detect meaningful interannual variation in key response variables? What sample size is required?

This is addressed using **paired-year comparisons**, simulation of a 20% change, and model-based estimation of detection probability and uncertainty. The analysis evaluates how many plots are needed to reliably detect moderate ecological change.

## Goal 3: Detecting Temporal Trends
> Can long-term temporal trends in species abundance be detected at each site? What sampling intensity and species characteristics affect trend detectability?

This is now addressed using **temporal modeling across all available years**, with `year` as a covariate in the GJAM model. It enables detection of increasing or decreasing trends over time for each species.

# Two-Part Modeling Framework

This pipeline implements both components:

### Part 1: Year-to-Year Change Detection
- Compare adjacent years
- Simulate 20% increase in one year
- Estimate posterior detection probability per species
- Assess power across subsampled plots

### Part 2: Long-Term Trend Detection
- Include all years of data
- Model year as continuous covariate (`~ year`)
- Estimate posterior slope per species
- Assess CI overlap with zero and relationship with abundance

# Part 1: Year-to-Year Change Results

```{r detection-summary-plot, echo=TRUE, message=FALSE, warning=FALSE}
det_summary <- readRDS("_targets/objects/detection_summary")
plot_detection_prob(det_summary)
```

```{r summary-table, echo=FALSE}
kable(det_summary, digits = 3, caption = "Detection Probabilities with Credible Intervals") %>%
  kable_styling(full_width = FALSE)
```

# Interpreting Change Detection Sensitivity

Posterior estimates allow evaluation of whether simulated changes are detectable under different sampling scenarios.

> A species is detectable under a 20% change if the posterior 95% credible interval for detection probability does not overlap with baseline, and the lower bound is > 0.5.

```{r sample-size-detection-summary, echo=TRUE, message=FALSE, warning=FALSE}
# Placeholder summarization by sample size
summarize_detection_by_size <- function(det_data) {
  det_data %>%
    group_by(sample_size) %>%
    summarise(
      n_species = n(),
      n_detectable = sum(mean > 0.5 & lower > 0.1),
      pct_detectable = n_detectable / n_species * 100
    )
}
```

# Part 2: Long-Term Trend Analysis

```{r trend-results-placeholder, echo=TRUE, message=FALSE, warning=FALSE}
trend_summary <- readRDS("_targets/objects/trend_summary")
head(trend_summary)
```

This section uses GJAM models fitted to full multi-year data with `year` as a covariate to assess trend detectability:
- Species with high posterior slope and tight CI indicate strong trends
- Thresholds for detectability can be plotted against abundance or frequency

```{r trend-plot-placeholder, echo=FALSE}
# Example placeholder (insert real plot code)
# ggplot(trend_summary, aes(x = abundance, y = slope, color = trend_detectable)) +
#   geom_point() + geom_errorbar(aes(ymin = slope_lower, ymax = slope_upper))
```

# Summary

This two-part analysis provides:
- **Plot-level sampling guidance** to detect moderate year-to-year change
- **Species-specific trend profiles** for multi-year ecological monitoring

Future work may include:
- Species grouping by growth form or abundance class
- Assessing climate or land use drivers of trends
- Publishing detection maps or early-warning metrics

---

*Developed with `targets`, `gjam`, and `ggplot2`.*
