---
title: "Detection Probability Report"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tibble)
```

## Overview

This report summarizes the detection probability estimates for plant species cover changes across NEON sites using GJAM-based simulations.

## Detection Summary Table

```{r}
# Load detection summary if not already available
if (!exists("detection_summary")) {
  detection_summary <- readRDS("outputs/detection_summary.rds")
}

head(detection_summary)
```

## Plot: Detection Probability by Species and Subsample

```{r}
if (!exists("detection_summary")) {
  detection_summary <- readRDS("outputs/detection_summary.rds")
}

# Basic plot by species and sample size
ggplot(detection_summary, aes(x = subsample_size, y = detection_prob, color = species)) +
  geom_line() +
  facet_wrap(~ siteID) +
  labs(
    title = "Detection Probability by Subsample Size",
    x = "Subsample Size (Number of Plots)",
    y = "Probability of Detecting 20% Change"
  ) +
  theme_minimal()
```

## Plot: Detection Probability Across Sites

```{r}
# Site-level average detection
detection_summary %>%
  group_by(siteID, subsample_size) %>%
  summarize(mean_prob = mean(detection_prob, na.rm = TRUE)) %>%
  ggplot(aes(x = subsample_size, y = mean_prob)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Detection Probability Across Sites",
    x = "Subsample Size",
    y = "Average Detection Probability"
  ) +
  theme_minimal()
```

## Interpretation and Decision Support

- This report can guide decisions on optimal sampling effort (number of plots) to detect 20% change in percent cover for species at NEON sites.
- Smaller sample sizes may be acceptable at sites with higher detection power.
- Consider additional simulations for edge cases or rare taxa.
