---
title: "NEON GJAM Detection Report (read-only)"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  pdf_document:
    number_sections: true
    fig_caption: true
fontsize: 11pt
geometry: margin=1in
params:
  OUTPUT_DIR: "outputs"
  example_site: "JERC"
  top_n_species: 6
  min_sample_size: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
suppressPackageStartupMessages({
  library(arrow); library(dplyr); library(ggplot2); library(tidyr); library(stringr)
})
`%||%` <- function(a, b) if (!is.null(a)) a else b

OUTPUT_DIR <- params$OUTPUT_DIR %||% Sys.getenv("OUTPUT_DIR", unset = "outputs")
sp_path <- file.path(OUTPUT_DIR, "species_summary.parquet")
dr_path <- file.path(OUTPUT_DIR, "draws_index.parquet")
rc_path <- file.path(OUTPUT_DIR, "relative_cover.parquet")

stopifnot(file.exists(sp_path), file.exists(dr_path), file.exists(rc_path))

species_summary <- arrow::read_parquet(sp_path)
draws_index     <- arrow::read_parquet(dr_path)
relative_cover  <- arrow::read_parquet(rc_path)

need <- function(df, cols, label) {
  miss <- setdiff(cols, names(df))
  if (length(miss)) {
    cat(sprintf("**Skipped %s** — missing columns: %s\n\n", label, paste(miss, collapse=", ")))
    return(FALSE)
  }
  TRUE
}

# simple figure counter (1,2,3...)
.fig_counter <- 0L
figcap <- function(txt) { .fig_counter <<- .fig_counter + 1L; paste0("Figure ", .fig_counter, ". ", txt) }

# parameters
threshold <- 0.8
pair_coverage_target <- 0.75
example_site <- params$example_site %||% (species_summary$site %>% unique() %>% sort() %>% .[1])
MIN_N <- params$min_sample_size %||% 5
```

# Methods (brief)

We fit a single joint species model (GJAM, continuous abundance) per site with predictors **year** (factor) and **nlcdClass** (factor). We compute species-level detection probabilities for a +20% change. This report is **read-only**: it assumes standardized inputs and performs only aggregation and plotting.

Community-weighted mean (CWM) definition:
\[
\mathrm{CWM} = \frac{\sum_s w_s D_s}{\sum_s w_s}
\]
with relative-cover weights \(w_s\) and species detection \(D_s\).

# Species-level detection (example site)

```{r species-level}
req <- c("site","species","year_pair","sample_size","replicate","mean_detection","mode")
if (need(species_summary, req, "Species-level detection")) {

  # Top species by relative cover at the example site
  top_species <- relative_cover %>%
    rename(site = siteID, species = taxonID) %>%
    filter(site == example_site) %>%
    group_by(species) %>%
    summarise(mean_rc = mean(relative_cover, na.rm = TRUE), .groups = "drop") %>%
    arrange(desc(mean_rc)) %>%
    slice_head(n = params$top_n_species) %>%
    pull(species)

  var_sum <- species_summary %>%
    filter(site == example_site, mode == "sensitivity_variable", species %in% top_species, sample_size >= MIN_N) %>%
    group_by(species, year_pair, sample_size) %>%
    summarise(
      mean_det = mean(mean_detection, na.rm = TRUE),
      lo = if ("ci_lower" %in% names(cur_data_all())) mean(ci_lower, na.rm = TRUE) else NA_real_,
      hi = if ("ci_upper" %in% names(cur_data_all())) mean(ci_upper, na.rm = TRUE) else NA_real_,
      .groups = "drop"
    )

  p <- ggplot(var_sum, aes(sample_size, mean_det, color = species, group = species)) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.3, alpha = 0.5) +
    geom_hline(yintercept = 0.8, linetype = 2) +
    scale_y_continuous(limits = c(0,1)) +
    facet_wrap(~ year_pair, scales = "free_x") +
    labs(x = "Sample size (plots)", y = "Detection probability",
         title = paste0("Species-level detection (", example_site, ")"))
  p
}
```

# Community-weighted detection (approximate, by year pair)

```{r cwm-by-pair}
req1 <- c("site","species","year_pair","sample_size","replicate","mean_detection","mode")
req2 <- c("siteID","taxonID","year","relative_cover")
if (need(species_summary, req1, "CWM by pair") && need(relative_cover, req2, "relative cover")) {

  rc_mean <- relative_cover %>%
    rename(site = siteID, species = taxonID) %>%
    filter(site == example_site) %>%
    group_by(species) %>%
    summarise(w = mean(relative_cover, na.rm = TRUE), .groups = "drop")

  # Approximate CWM from species-level means (per pair × N)
  rep_avg_var <- species_summary %>%
    filter(site == example_site, mode == "sensitivity_variable", sample_size >= MIN_N) %>%
    left_join(rc_mean, by = "species") %>%
    filter(!is.na(w)) %>%
    group_by(year_pair, sample_size, replicate) %>%
    summarise(mean_detection = sum(mean_detection * w) / sum(w), .groups = "drop") %>%
    group_by(year_pair, sample_size) %>%
    summarise(mean_detection = mean(mean_detection, na.rm = TRUE), .groups = "drop")

  p <- ggplot(rep_avg_var, aes(x = sample_size, y = mean_detection, color = year_pair, group = year_pair)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    geom_hline(yintercept = threshold, linetype = "dashed") +
    scale_y_continuous(limits = c(0,1)) +
    scale_x_continuous(breaks = sort(unique(rep_avg_var$sample_size))) +
    labs(x = "Sample size (plots)", y = "CWM detection", color = "Year pair",
         title = paste0("Community-weighted detection (", example_site, ")"))
  p
}
```

# Mean across year pairs (example site)

```{r cwm-mean}
req1 <- c("site","species","year_pair","sample_size","replicate","mean_detection","mode")
req2 <- c("siteID","taxonID","year","relative_cover")
if (need(species_summary, req1, "CWM mean") && need(relative_cover, req2, "relative cover")) {

  rc_mean <- relative_cover %>%
    rename(site = siteID, species = taxonID) %>%
    filter(site == example_site) %>%
    group_by(species) %>%
    summarise(w = mean(relative_cover, na.rm = TRUE), .groups = "drop")

  site_avg_var <- species_summary %>%
    filter(site == example_site, mode == "sensitivity_variable", sample_size >= MIN_N) %>%
    left_join(rc_mean, by = "species") %>%
    filter(!is.na(w)) %>%
    group_by(year_pair, sample_size, replicate) %>%
    summarise(cwm = sum(mean_detection * w) / sum(w), .groups = "drop") %>%
    group_by(sample_size) %>%
    summarise(detection_mean = mean(cwm, na.rm = TRUE), .groups = "drop")

  ggplot(site_avg_var, aes(x = sample_size, y = detection_mean, group = 1)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2) +
    geom_hline(yintercept = threshold, linetype = "dashed") +
    scale_y_continuous(limits = c(0,1)) +
    scale_x_continuous(breaks = sort(unique(site_avg_var$sample_size))) +
    labs(x = "Sample size (plots)", y = "CWM detection",
         title = paste0("Mean across year pairs (", example_site, ")"))
}
```

# Decision metric: fraction of year pairs ≥ 0.8 (all sites)

```{r decision-metric}
req <- c("site","year_pair","sample_size","replicate","mean_detection","mode")
if (need(species_summary, req, "Decision metric")) {

  # For simplicity, use species-level mean_detection averaged within pair × N (unweighted)
  pair_means <- species_summary %>%
    filter(mode == "sensitivity_variable") %>%
    group_by(site, year_pair, sample_size, replicate) %>%
    summarise(md = mean(mean_detection, na.rm = TRUE), .groups = "drop") %>%
    group_by(site, year_pair, sample_size) %>%
    summarise(md = mean(md, na.rm = TRUE), .groups = "drop")

  frac_tbl <- pair_means %>%
    mutate(hit = md >= threshold) %>%
    group_by(site, sample_size) %>%
    summarise(frac_pairs = mean(hit), .groups = "drop")

  ggplot(frac_tbl, aes(x = sample_size, y = frac_pairs, group = site)) +
    geom_line(alpha = 0.5) +
    geom_point(alpha = 0.7, size = 1.8) +
    scale_y_continuous(limits = c(0,1)) +
    scale_x_continuous(breaks = sort(unique(frac_tbl$sample_size))) +
    labs(x = "Sample size (plots)", y = "Fraction of year pairs ≥ 0.8",
         title = "Decision metric by site")
}
```

# Posterior draw illustration (optional)

```{r posterior-illustration}
# This section uses a single shard path if a 'file' column is present in draws_index.
if ("file" %in% names(draws_index)) {
  d1 <- draws_index %>%
    filter(sample_size >= MIN_N) %>%
    arrange(sample_size)
  if (nrow(d1)) {
    fp <- d1$file[[1]]
    if (is.character(fp) && file.exists(fp)) {
      dd <- tryCatch(arrow::read_parquet(fp), error = function(e) NULL)
      if (!is.null(dd) && "detected" %in% names(dd)) {
        ggplot(dd, aes(x = as.numeric(detected))) +
          geom_histogram(bins = 2) +
          labs(x = "Detected (0/1)", y = "Count", title = "Example draws shard")
      } else {
        cat("**Skipped** — draws shard missing or no `detected` column.\n")
      }
    } else {
      cat("**Skipped** — shard path not found.\n")
    }
  } else {
    cat("**Skipped** — no draws index rows.\n")
  }
} else {
  cat("**Skipped** — `file` column not present in draws_index.\n")
}
```
