---
title: "&nbsp;"
fontsize: 11pt
geometry: margin=1in, includefoot, top=1in, bottom=1.25in
header-includes:
- \usepackage{fancyhdr}
- \usepackage{ragged2e}
#- \usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
- \usepackage{lastpage}
- \usepackage{hyperref}
- \usepackage[hypcap]{caption}
- \usepackage{array}
- \usepackage{graphicx}
- \usepackage[datesep=/]{datetime2}
- \usepackage{ltablex}
- \usepackage{pifont}
- \usepackage{float}
- \usepackage{longtable}
- \setlength{\headsep}{0.7in}
linkcolor: black
mainfont: Calibri
output:
  pdf_document:
    fig_caption: yes
    fig_width: 4
    latex_engine: xelatex
    number_sections: yes
  html_document: default
  word_document: default
---
```{r libraries, echo=FALSE, message=FALSE, warning=FALSE}

#load libraries
#if packages listed below are not installed, use install.packages("packageName") before running library(packageName)# 

library(knitr)
library(kfigr)
library(dplyr)
library(stringr)
library(readr)


# install TeX program :  http://miktex.org/2.9/setup (Windows) ; http://tug.org/mactex/ (Mac)
# re-install, update miktex - datetime2 package (called here) was released 3/24/15, all older versions do not have this package pre-loaded.
# documentation for kfigr https://github.com/mkoohafkan/kfigr/blob/master/vignettes/introduction.Rmd

```

```{r titlePage, echo=FALSE}

#Change for each document

#Title
title <- "TOS Design Optimization: Plant Diversity Sampling" #change AOS/TOS and Protocol Name

#Authors
firstAuthor <- "Dave Barnett"
firstAuthorDate <- "03/03/2026"
firstAuthorOrg <- "SCI"

secondAuthor <- "Eric Sokole"
secondAuthorDate <- "03/03/2026"  
secondOrg <- "SCI"

thirdAuthor <- "Courtney Meier"
thirdAuthorDate <- "03/03/2026"
thirdOrg <- "SCI"

#repeat or delete as necessary 

#Approvals
approvalName <- "Kate Thibault"
approvalOrg <- "SCI"
approvalDate <- "mm/dd/yyyy"

secondApprovalName <- "Approver 2"
secondApprovalOrg <- "SYS"
secondApprovalDate <- "mm/dd/yyyy"

#repeat or delete as necessary 

#Release
releasedByName <- "Releaser"
releasedByOrg <- "CM"
releasedByDate <- "mm/dd/yyyy"

```

[//]: TEMPLATE SECTION 1 - DO NOT CHANGE!! #####################################
[//]: These are document settings for NEON template, code will not print


\DTMsetdatestyle{mmddyyyy}
\thispagestyle{fancy}
\sloppy
\RaggedRight
\pagestyle{fancy}
\pagenumbering{gobble}
\fancyhf{}
\fancyheadoffset[L, R]{0.5in}
\fancyfootoffset[L, R]{0.5in}
\fancyhead[L]{\includegraphics[width=4cm]{Logo}
\vspace{0.5cm}} 

\fancyhead[R]{
\footnotesize
\setlength{\extrarowheight}{1mm}
\begin{tabular}{|p{5cm}|p{5cm}|p{1in}|}
\hline
\multicolumn{2}{|p{11cm}|}{\emph{Title:} `r title`} & \emph{Date:} `r firstAuthorDate` \\ \hline
\emph{}    & \emph{Author:} `r firstAuthor`        & \emph{}  \\ \hline
\end{tabular}
\vspace{1cm}
} 
        
[//]: check out http://www.tablesgenerator.com/latex_tables for help with code to generate tables

\renewcommand{\headrulewidth}{0pt}
\maketitle
\thispagestyle{fancy}

\begin{center}

\vspace{0.3in}
\huge \bfseries \uppercase {`r title` \\[0.4cm] }

\end{center}

\vspace{1in}

\setlength{\extrarowheight}{1mm}
\begin{tabular}{|m{5.5cm}|m{5.5cm}|m{5.5cm}|}
\hline
\textbf{PREPARED BY} &\textbf{ORGANIZATION}  &\textbf{DATE} \\ \hline
`r firstAuthor`      & `r firstAuthorOrg`         & `r firstAuthorDate` \\ \hline
`r secondAuthor`     & `r secondOrg`         & `r secondAuthorDate` \\ \hline
`r thirdAuthor`     & `r thirdOrg`         & `r thirdAuthorDate` \\ \hline
\end{tabular}


\setlength{\extrarowheight}{1mm}
\begin{tabular}{|m{5.5cm}|m{5.5cm}|m{5.5cm}|}
\hline
\textbf{APPROVALS}     &\textbf{ORGANIZATION}  &\textbf{APPROVAL DATE} \\ \hline
`r approvalName`       & `r approvalOrg`       &  `r approvalDate`\\ \hline
`r secondApprovalName` & `r secondApprovalOrg` &`r secondApprovalDate`  \\ \hline
\end{tabular}

\setlength{\extrarowheight}{1mm}
\begin{tabular}{|m{5.5cm}|m{5.5cm}|m{5.5cm}|}
\hline
\textbf{RELEASED BY} &\textbf{ORGANIZATION}  &\textbf{RELEASE DATE} \\ \hline
`r releasedByName` & `r releasedByOrg`        &  `r releasedByDate` \\ \hline
\end{tabular}

\vspace{1.5in}

\begin{center}
See configuration management system for approval history.

\bigskip



\scriptsize{The National Ecological Observatory Network is a project solely funded by the National Science Foundation and managed under cooperative agreement by Battelle. Any opinions, findings, and conclusions or recommendations expressed in this material are those of the author(s) and do not necessarily reflect the views of the National Science Foundation.}
\end{center}


\fancyfoot[L]{
\footnotesize{\color{gray}Template\_NEON.DOC.004243 Rev J}
} 

\newpage

\vspace{3in}


\newpage
\pagenumbering{roman}
\fancyfoot[C]{\thepage}
\fancyfoot[L]{
\footnotesize{}
} 

\renewcommand\contentsname{\uppercase{Table of contents}}
\tableofcontents
\fancyfoot[C]{\thepage}
\renewcommand\listtablename{\uppercase{list of tables and figures}}
\let\oldnumberline\numberline
\renewcommand{\numberline}{\tablename~\oldnumberline}
\listoftables
\fancyfoot[C]{\thepage}
\renewcommand{\numberline}{\figurename~\oldnumberline}
\renewcommand\listfigurename{}
\addtocontents{lof}{\vskip -1.2cm}
\listoffigures
\fancyfoot[C]{\thepage}


\newpage


\pagenumbering{arabic} 
\fancyfoot[C]{
Page \thepage \hspace{1pt} of \pageref{LastPage}
}
<!---
This is super secret text that will not appear in the pdf. 

The \changeMe command turns text red to indicate that section should be modified for each document. Remove the \changeMe{} to make the text black
--->

\newcommand{\changeMe}[1]{\textcolor{red}{#1}}

# BACKGROUND AND GOALS

The National Ecological Observatory Network (NEON) generates data and samples to advance understanding of ecological processes and inform management of US ecosystems. Multiple tiers of requirements provide traceable links between this high-level NEON mission statement and the protocols that direct data collection. The Terrestrial Observation System (TOS) science design for plant diversity ([RD03]) and the resulting study design, protocol ([RD03]), and data product (Plant Presenece and Percent Cover, DP1.10058.001) are designed to enable status and trends in composition, abundance, and richness of plant species and function types at multiple scales. 

## Description of sample design and available data

NEON's plot-based sampling of plant species diversity satisfies the requirements that "NEON's measurement strategy will include biological invaders and diversity of key organismal groups" and "NEON measurements of plant diversity will include native and invasive plant species (Schimel et al. 2011)." Species are observed with the Carolina Vegetation Survey (Peet et al. 1998) multi-scale plot design that borrows from a sampling approach pioneered by Robert Whittaker (Shmida 1984, Stohlgren 1995). NEON samples with a 20 x 20m square plot with four 10m x 10m subplots containing nested 10m$^2$ and 1m$^2$ subplots (Figure 1). Specifically: 

*  The identity of each species (or the lowest taxonomic rank possible) according to naming conventions maintained by the US Department of Agriculture, Natural Resources Conservation Service PLANTS database (USDA, NRCS 2016) is recorded in each subplot - eight 1m$^2$, eight 10m$^2$, four 100m$^2$. 
*  Estimates of taxon-specific abundance is made with estimates of cover within the 1m$^2$ subplots. 

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth,height=\textheight,keepaspectratio]{plotLayout}
\caption{A diagram of the multiscale plot for observations of plant cover-abundance and occurrence data.  }
\end{figure}

At each of 47 terrestrial NEON sites, the NEON spatial sampling design for organisms and soil directs observations to (Thorpe et al. 2016):

*  Tower Base Plots: Randomly placed plots near the NEON flux tower designed to calibrate sensor-based measures of atmosphere and soil  
*  Distributed Base Plots: Plots to measure variability across larger landscapes according to a stratified-random design  

The plant diversity protocol is implemented at three of the 20–30 Tower Base Plots. The majority of the plant species diversity sampling effort occurs at the Distributed Base Plots (Thorpe et al. 2016). Ecological literature (e.g., Stohlgren 2007), early input from the ecological community, and power analyses (Thompson et al. 2012, Barnett et al. 2018) resulted in an initial sample size of 20–30 Distributed Base plots at each site.

The frequency and timing of sampling is guided by NEON requirements (Schimel et al. 2011). Implementing the protocol at each plot at least one time each year enables the quantification of annual rates of change. These sampling bouts occur during a one to two month period when the majority of species flower or possess other parts conducive to identification. This sampling window targets annual peaks in greenness as measured by MODIS, and incorporates input from local ecologists and NEON phenology observations (Meier et al. 2023). The few NEON sites characterized by distinct biomodal patterns in peak greenness populated disparate suites of apparent species are sampled twice annually.  For example, the Sonoran desert (represented by the NEON site at the Santa Rita Experimental Range) gets rain in late winter and a monsoonal moisture pulse late in the summer. Plant species adapted to the particular intersection of temperature and the timing and amount of precipitation grow and flower and can be identified only during one of the two distinct events (Ogle and Reynolds 2004).

For the purpose of these analyses, data includes all available Distributed Base Plots from all 47 NEON sites (NEON 2025) downloaded directly from the application programming interface (API) and stacked with ‘neonUtilities::stackByTable()‘ function on 03/01/2026.  


## Analytical Goals

Statistically rigorous analyses are needed to assess the capacity of NEON data to address Observatory goals and to guide iterative sampling design optimization efforts. Initial spatial and temporal sampling designs for the NEON Terrestrial Observation System (TOS) and Aquatic Observation System (AOS) were developed in collaboration with Technical Working Groups (TWGs) comprised of community experts, and were captured in Science Design documents (RD[01]). The initial designs relied on analysis of published datasets (where relevant), analysis of NEON prototype data collection efforts, and subject matter expertise.  

As the NEON Observatory matures, it is critical for initial design assumptions to be tested with multiple years of data collected from NEON sites. Analyses must: 

*  Describe where the data support reduction of sampling effort without compromising NEON's ability to meet design requirements (e.g., target sites for reduction in sampling effort, spatial scales where replication is not necessary, etc.). 
*  Evaluate effects of design modifications on uncertainty and ability to detect changes for key response variables (e.g., effects of reduced spatial and/or temporal sampling effort). 

Analysis and evaluation of the data provides a feedback loop that enables assessment of the Designs relative to Observatory goals. Moreover, results of these analyses will allow the NEON to effectively prioritize sampling costs in the face of uncertain future funding levels.  

Specification of analyses suitable for evaluating the efficacy of the NEON sampling designs required development of specific questions. Power analyses and sampling simulations were then used to determine the degree to which design modifications affected the ability of the data to address those questions. Questions and analytical approaches were developed in consultation with NEON Technical Working Groups comprised of subject matter experts selected from the ecological community. 

The two primary and actionable questions of a 2018 effort for these data were:

\begin{enumerate}
   \item Where is the variability in the data? Specifically, what proportion of the variance at each site exists at the year, bout, plot, and subplot scales?  Understanding the components of variance provides a critical component of the actionable directions for modifying sampling size, subplot sampling intensity, and temporal sampling frequency. 
   
   \item Is it possible to detect meaningful interannual variation in key response variables? Understanding the power to detect change between years provides a direct indication of the sample sizes required to detect meaningful ecological change. 
\end{enumerate}

The variance components analysis supported the reduction in the number of the number nested 10m$^2$ and 1m$^2$ subplots within each plot from eight to six. Additionally, the 10m$^2$ and 100m$^2$ subplots at which species presence are recorded were reduced to every other year sampling. The 1m$^2$ subplots at each plot continue to be sampled annually.

The power analysis reduced the total number of plots sampled at a small number of sites. Only sites with three or more years of data were considered. An initial goal was to evaluate trend detection, but it was decided insufficient years of data had been collected for that analysis.

With many more years of data available across all NEON terrestrial sites in 2026 (Appendix A), this effort revisits the question of detectability of interannual change in plant community composition and abundance. Specifically, we ask: 

\begin{enumerate}
\item What level of spatial replication (number of plots) is required to reliably detect a specified relative change in plant species cover at a site?
\item What level of spatial replication is required for reliable estimation of multi-year directional trends in species cover?
\end{enumerate}

Together, these analyses provide complementary bounds on the minimum sampling effort required to detect meaningful ecological change at short (interannual) and long (multi-year trend) temporal horizons.


# RELATED DOCUMENTS AND ACRONYMS

##  Reference Documents

```{r Reference Document list, echo=FALSE}

# Documents will be assigned an RD[##] in the order they are listed

RDlist <- as.data.frame(rbind(
  "name1" = c("name"="NEON Observatory Design", "doc"= "NEON.DOC.000001"),
  "name2" = c("name"="NEON Level 1, Level 2, and Level 3 Data Products Catalog", "doc"= "NEON.DOC.002652"),
  "name3" = c("name"="TOS Science Design for Plant Diversity", "doc"="NEON.DOC.000912"),
  "name4" = c("name"="TOS Protocol and Procedure: Plant Diversity Sampling", "doc"="NEON.DOC.014042")
  # any additional documents...
))

RDlist <- cbind(RDlist,
  "ref"= paste("RD[",
              formatC(1:nrow(RDlist), width=2, flag="0"),
              "]", 
              sep="")
)

```
\begin{tabular}{|m{1cm}|m{3cm}|m{11cm}|}
\hline
`r paste(RDlist$ref, "&",
        RDlist$doc, "&",
        RDlist$name, "\\\\ ","\\hline", sep="", collapse=" ")`
\end{tabular}


##  Acronyms
<!---
Fill in the Acronym before the "&"" and the definition after the "&"
--->
\begin{tabular}{|m{3cm}|m{12.5cm}|}
\hline
\bfseries{Acronym}&  \bfseries{Definition}  \\ \hline
  NEON  &  National Ecological Observatory Network \\ \hline
  TOS   &  Terrestrial Observation System \\ \hline
 TWG   &  Technical Working Group \\ \hline
 NLCD  & National Land Cover Database \\ \hline
 GJAM  & Generalized Joint Attribute Model \\ \hline
 CA    & Censored Abundance (GJAM likelihood for cover data) \\ \hline
 MCMC  & Markov Chain Monte Carlo \\ \hline
\end{tabular}

# METHODS

We evaluated the capacity of NEON’s terrestrial plant sampling design to detect ecological change in species- and community-level response variables using a hierarchical Bayesian modeling framework combined with simulation-based power and sensitivity analyses. The assessment comprises two complementary analyses: 
\begin{enumerate}
   \item We conducted a perturbation-based power analysis to quantify the spatial replication (number of randomly placed plots) required to detect a specified magnitude of interannual change in species cover, evaluated through a paired counterfactual design that controls for natural year-to-year variability. 

  \item We evaluated the spatial replication required for reliable estimation of multi-year directional trends, including the stability of trend estimates across sample sizes and the detectability of imposed trends of known magnitude. 
\end{enumerate}

Together, these analyses address the two primary temporal scales at which change detection is relevant to NEON’s mission: abrupt interannual shifts and gradual directional trends.


## Data and preprocessing

The observation unit for modeling is a species-specific cover value at the plot-year level. Within each plot-year, the six (or eight prior to 2019) 1m$^2$ subplot observations are averaged to produce a single cover value per species per plot per year. For example, at a site where 30 plots are sampled annually for 8 years, a species recorded on all plots would contribute 240 plot-year observations to the model. The full model input for a site is the matrix of all species × all plot-years, with cover values as entries. 

Prior to model fitting, species occurring in fewer than three plots across the entire site record are aggregated into a composite "OTHER" category. This filter removes species for which spatial replication is insufficient to estimate covariance structure, while retaining them in the modeled community so that their aggregate contribution to total cover is represented (Clark et al. 2017). Species with all-zero cover or zero variance after aggregation are removed. The "OTHER" category participates in model fitting but is excluded from power and trend analyses, which focus on individually identifiable species.

Separately, a cover floor of 3\% mean raw field cover is applied when evaluating power and trend stability. Species below this threshold are excluded from the power analysis because a 20\% relative change at 3\% cover corresponds to 0.6 percentage points---below the 1\% recording resolution of NEON's field protocol. This cover floor is computed from the raw field observations (mean of \texttt{percent cover} per species across all plot-years at the site) and is independent of the GJAM model predictions which undergo compositional normalization that compresses cover values.


## Joint species distribution model

Both analyses share a common modeling framework. Plant species cover was modeled with generalized joint attribute models (GJAM; Clark et al. 2017), a multivariate hierarchical Bayesian framework that jointly represents all species at a site while accounting for shared responses to environmental covariates and residual covariance among species. A single GJAM model is fit per site using all available plot-years. Both the perturbation-based and trend analyses draw on the posterior output of this shared fit. For plot $p$ in year $t$, the vector of observed percent cover values for all $S$ species is modeled as:

$$
\mathbf{y}_{p,t} \sim \text{CA}(\boldsymbol{\mu}_{p,t}, \boldsymbol{\Sigma})
$$

$$
\boldsymbol{\mu}_{p,t} = \mathbf{X}_{p,t}\mathbf{B}
$$

where $\mathbf{y}_{p,t}$ is the $S$-vector of observed percent cover, $\boldsymbol{\mu}_{p,t}$ is the latent mean cover vector, $\mathbf{X}_{p,t}$ is the design matrix of covariates, $\mathbf{B}$ is the $K \times S$ matrix of species-specific regression coefficients (equivalent to the transpose of the coefficient matrix in Clark et al. 2017), and $\boldsymbol{\Sigma}$ represents residual covariance among species not explained by the covariates, where $K$ is the number of covariates.

The censored-abundance (CA) likelihood represents observed cover as a censored realization of an underlying multivariate normal latent variable, allowing exact zeros as well as continuous positive values.

Two model specifications were used depending on land cover heterogeneity at each site: $\mathbf{Y} \sim \text{year} + \text{nlcdClass}$ at sites with multiple National Land Cover Database (NLCD) classes represented among sample plots, and $\mathbf{Y} \sim \text{year}$ at sites with a single class or where the multi-class specification failed to converge. Year was treated as a categorical factor. The MCMC sampler was run for 5,000 iterations; 2,500 iterations were discarded as burn-in, yielding 2,500 retained posterior draws. For downstream posterior prediction and sensitivity analyses, we used $G = 1{,}000$ posterior draws sampled from the retained MCMC output for computational efficiency.


### Posterior prediction and dimension reduction

Joint modeling of plant communities at NEON sites frequently involves hundreds of species observed across tens of plots(Appendix here to numbers of sp), producing covariance matrices that are high-dimensional relative to the available sample size. To illustrate, consider a southeastern forest site with 400 plant species observed across 30 plots per year: estimating the $400 \times 400$ residual covariance matrix $\boldsymbol{\Sigma}$ from 300 plot-year observations is not feasible using standard approaches.

GJAM addresses this challenge through a Bayesian dimension reduction framework (Taylor-Rodriguez et al. 2017; Clark et al. 2017). Rather than estimating $\boldsymbol{\Sigma}$ directly, the model represents the $S$-dimensional species response through a lower-dimensional factor structure. Species are assigned to $N_c$ clusters via a truncated Dirichlet process representation, and within each cluster, an $r$-dimensional factor loading matrix $\mathbf{Z}$ captures shared covariance. The implied covariance is:

$$
\boldsymbol{\Sigma} = \mathbf{A}\mathbf{A}' + \sigma^2_\varepsilon \mathbf{I}
$$

where $\mathbf{A} = \mathbf{Q}(\mathbf{k})\mathbf{Z}$ is the $S \times r$ loading matrix that maps the reduced factor structure back to the full species space via cluster assignments $\mathbf{k}$, and $\sigma^2_\varepsilon$ is a scalar residual variance. This factorization reduces the effective number of covariance parameters from $S(S+1)/2$ to $(N_c \times r) + 1$, enabling stable estimation even when $S$ greatly exceeds the number of observations $n$.

GJAM invokes dimension reduction when the number of species is large relative to the number of observations, which was the case for all NEON sites analyzed here (Taylor-Rodriguez et al. 2017). This condition was met at all 46 NEON sites, where species counts ranged from 39 to 406 while plot-year observations ranged from approximately 100 to 400. We therefore set reduction explicitly with a truncation level of $N_c = 20$ clusters and $r = 5$ factors across all sites, ensuring consistent parameterization and comparability of posterior predictions across the Observatory. Posterior predictive draws were reconstructed from the stored MCMC chains for $\mathbf{B}$, $\mathbf{k}$, $\mathbf{Z}$, and $\sigma^2_\varepsilon$ using the reduced-rank generative model:

$$
\mathbf{y}_i^{(g)} = \mathbf{x}_i \mathbf{B}^{(g)} + \mathbf{w}_i^{(g)} {\mathbf{A}^{(g)}}' + \boldsymbol{\varepsilon}_i^{(g)}
$$

where $\mathbf{x}_i \mathbf{B}^{(g)} = \boldsymbol{\mu}_i^{(g)}$ is the latent mean response defined above, $\mathbf{B}^{(g)}$ is the $K \times S$ matrix of species-specific regression coefficients for posterior draw $g$, $\mathbf{A}^{(g)}$ is the $S \times r$ loading matrix reconstructed from cluster assignments $\mathbf{k}^{(g)}$ and factor loadings $\mathbf{Z}^{(g)}$ at draw $g$, $\mathbf{w}_i^{(g)}$ is an $r$-dimensional latent factor score with $\mathbf{w}_i^{(g)} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_r)$, and $\boldsymbol{\varepsilon}_i^{(g)}$ is species-level residual error with $\boldsymbol{\varepsilon}_i^{(g)} \sim \mathcal{N}(\mathbf{0}, \sigma^{2(g)}_\varepsilon \mathbf{I})$. Superscript $(g)$ indexes the posterior draw used for prediction ($g = 1, \ldots, G$, with $G = 1{,}000$ in the sensitivity analyses). For each draw, the parameters $\mathbf{B}^{(g)}$, $\mathbf{A}^{(g)}$, and $\sigma^{2(g)}_\varepsilon$ are fixed, while $\mathbf{w}_i^{(g)}$ and $\boldsymbol{\varepsilon}_i^{(g)}$ are drawn independently, yielding posterior predictive realizations that preserve the covariance structure implied by the reduced-rank model without explicitly forming the full $S \times S$ covariance matrix.

The analysis propagates key sources of uncertainty through the posterior predictive distribution. Parameter uncertainty is captured through variation across MCMC draws, while residual variability—representing ecological and measurement variation not explained by covariates—is captured through the stochastic components $\mathbf{w}$ and $\boldsymbol{\varepsilon}$. Observation uncertainty (field measurement error) is not modeled explicitly but is implicitly absorbed into the residual variance estimates.

## Analytical framework

Both analyses use a common modeling and prediction framework to evaluate the detectability of ecological change. Posterior predictive draws from a single site-level model provide a representation of expected species abundance under observed environmental and spatial variability. These predictions are then combined with controlled perturbations and plot-level subsampling to quantify how sampling intensity influences inference.

The two analyses differ only in the temporal structure of the signal being evaluated. The counterfactual analysis focuses on discrete changes between paired states, while the trend analysis evaluates directional change accumulated over multiple years. In both cases, inference is based on the behavior of posterior predictive summaries under repeated subsampling of plots, allowing direct comparison of how sampling design affects the ability to detect and estimate change.


## Year-to-year change detection (counterfactual analysis)

The objective of the counterfactual analysis is to evaluate the extent to which changes in species abundance can be detected given the spatial variability observed within NEON sites. Because plant communities are spatially heterogeneous, the detectability of change depends not only on the magnitude of the signal, but also on the number and configuration of plots used to estimate site-level means.

To address this, we used posterior predictive draws from a single site-level model to generate paired baseline and perturbed realizations of species cover. The perturbation represents a controlled, biologically meaningful change applied to the latent mean structure, allowing direct comparison of predicted site-level responses under identical underlying variability.

For each site, detection was evaluated as a function of the number of plots sampled per year ($K$). Subsampling plots introduces variability analogous to field sampling, allowing the analysis to quantify how sampling intensity influences the probability of detecting a specified change. The result is a set of detection curves that describe the relationship between sampling effort and the probability of detecting a change of magnitude $\delta$.

This framework isolates the role of spatial sampling in determining detectability. Because the baseline and perturbed predictions are derived from the same posterior draws, differences between them reflect only the imposed change and the variability introduced by subsampling, rather than differences in model fit or parameter uncertainty.

This analysis was motivated by earlier approaches that evaluate change between consecutive years, but extends that framework by removing the confounding influence of interannual ecological variability. Rather than comparing observations from different years, both baseline and perturbed predictions are generated from the same underlying data and posterior parameter draws, with independent realizations of observation-level variability ($\mathbf{w}$ and $\boldsymbol{\varepsilon}$). This counterfactual design isolates statistical detectability from ecological variability among years, providing a controlled evaluation of monitoring performance under realistic covariance structure.

Let $\mu_{s,p}^{(g)}$ denote the posterior predictive cover for species $s$ at plot $p$ for MCMC draw $g$.

The perturbation was applied directly to the posterior predictive draws as a per-species additive shift proportional to each species' mean raw field cover:

$$
\mu_{s,p}^{\text{pert},(g)} = \mu_{s,p}^{(g)} + \delta \cdot \bar{c}_s
$$

where $\mu_{s,p}^{(g)}$ is the posterior predictive cover for species $s$ at plot $p$ for draw $g$, $\mu_{s,p}^{\text{pert},(g)}$ is the corresponding perturbed value, $\bar{c}_s$ is the mean percent cover of species $s$ computed from raw field observations across all plot-years at the site, and $\delta$ is the relative effect size. A species at 10\% mean cover with $\delta = 0.20$ receives a +2 percentage point perturbation; a species at 60\% mean cover receives +12 percentage points. This per-species scaling ensures every species faces the same proportional challenge regardless of baseline cover level. We tested effect sizes of $\delta \in \{0.15, 0.20, 0.25, 0.30\}$, corresponding to 15--30\% relative changes in cover.

For each site and year, the baseline and perturbed prediction sets were subsampled to represent alternative levels of spatial replication. All plots sampled in a given year were eligible for subsampling; no requirement was imposed that plots be sampled in consecutive years, as the paired counterfactual design generates both prediction sets from the same year's data. Subsamples consisted of $K$ plots drawn without replacement, with $K \in \{30, 25, 20, 15, 10, 8, 6, 5\}$, and each subsampling was repeated 50 times.

For each Monte Carlo replicate, the site-level mean cover for species $s$ was computed for both the baseline and perturbed prediction sets by averaging across the $K$ sampled plots within each posterior draw $g$:

$$
\bar{\mu}_{s,K}^{(g)} = \frac{1}{K}\sum_{p=1}^{K} \mu_{s,p}^{(g)}, 
\quad
\bar{\mu}_{s,K}^{\text{pert},(g)} = \frac{1}{K}\sum_{p=1}^{K} \mu_{s,p}^{\text{pert},(g)}
$$

where $\mu_{s,p}^{(g)}$ and $\mu_{s,p}^{\text{pert},(g)}$ denote the baseline and perturbed posterior predictive cover values, respectively.

Detection was evaluated using a per-species absolute difference criterion. For each species $s$, detection probability was defined as:

$$
P\left(\left|\bar{\mu}_{s,K}^{\text{pert},(g)} - \bar{\mu}_{s,K}^{\text{base},(g)}\right| \ge \tau_s\right) \ge 0.80
$$

where $\bar{\mu}_{s,K}^{\text{base},(g)}$ and $\bar{\mu}_{s,K}^{\text{pert},(g)}$ are the site-level mean cover values computed from the baseline and perturbed prediction sets, respectively, for posterior draw $g$, and $\tau_s$ is the species-specific detection threshold. The probability is computed as the fraction of posterior draws $g$ for which the absolute difference exceeds $\tau_s$. The ratio of imposed effect to detection threshold is constant across species (2:1 at $\delta = 0.20$), providing equivalent statistical difficulty regardless of cover level. This ensures that detectability comparisons are not driven by differences in species abundance scale.

A species was classified as detectable at sample size $K$ and effect $\delta$ when the above criterion was satisfied in at least 80\% of the 50 Monte Carlo replicates (power $\ge$ 0.80). The community-level metric reports the percentage of species (above the 3\% cover floor) that are individually detectable at each $K$.


## Multi-year trend analysis

### Objective and conceptual framing

Where the counterfactual analysis evaluates the detectability of discrete changes, the trend analysis addresses a complementary question: whether directional change in species abundance can be reliably estimated over time. In this context, the focus shifts from detecting a difference between two states to estimating the rate of change across multiple years.

As with the counterfactual analysis, inference is based on posterior predictive draws from a single site-level model. For each species, predicted cover values are generated for all plot–year combinations, preserving the spatial and temporal structure of the observed data. These predictions are then used to estimate annual site-level means under varying levels of spatial replication.

To evaluate the influence of sampling intensity, plots are subsampled independently within each year, and site-level mean cover is computed for each posterior draw. A linear trend is then estimated across years for each species, yielding a posterior distribution of slope parameters that reflects both model uncertainty and the variability introduced by spatial subsampling.

Two complementary properties of the estimated trend are evaluated. First, detection assesses whether the magnitude of the estimated slope exceeds a specified threshold, analogous to the threshold-based detection used in the counterfactual analysis. Second, stability evaluates the sensitivity of the estimated slope to the selection of plots, quantified as the coefficient of variation of slope estimates across repeated subsamples at each sample size.

Together, these metrics describe both the ability to detect directional change and the reliability of the estimated rate of change. As in the counterfactual analysis, the objective is not to optimize model fit, but to quantify how spatial sampling constrains inference about ecological change.

Many ecological changes of interest to NEON manifest as gradual directional trends that accumulate over multiple years. A site may experience a slow decline in native grass cover or a progressive increase in woody encroachment that is not detectable in any single year-to-year comparison but becomes apparent over a longer observation window. This analysis addresses the question of what level of spatial replication is required for reliable estimation and detection of such trends.

This question was identified as a priority during the 2018 NEON design optimization effort (Barnett et al. 2018) but could not be addressed at that time due to limited temporal replication. With 8--12 years of data now available at most NEON sites, the question is now tractable.


### Moving-window framework

Trend detection was evaluated using a moving-window approach (Urquhart et al. 1998). Rather than fitting a single trend line through the entire time series, we applied overlapping windows of fixed width $W$ that slide across the available years. This approach reflects realistic monitoring practice: an analyst examining NEON data in any given year would assess the most recent $W$ years to determine whether a trend is present. It also accommodates the possibility that trend strength varies across the observation period.

For each site, we evaluated window widths of $W \in \{5, 10\}$ years. Within each window, a trend was quantified as the ordinary least-squares slope of species cover against year, computed from posterior predictive site-year means derived from plot-level predictions. This yields a posterior distribution of slope estimates for each species, window position, and sample size.

<!-- FIGURE PLACEHOLDER -->
<!-- Conceptual figure: moving window illustration showing overlapping 5-year and 10-year windows sliding across a 10-year time series -->


### Observed trend detection

The observed trend detectability analysis evaluates whether trends already present in the posterior predictions are detectable as sample size changes. For each species, window position, posterior draw, and sample size $K$, the site-level mean cover was computed by averaging across the $K$ sampled plots within each year:

$$
\bar{\mu}_{s,t,K}^{(g)} = \frac{1}{K} \sum_{p=1}^{K} \mu_{s,p,t}^{(g)}
$$

where $\mu_{s,p,t}^{(g)}$ is the posterior predictive cover for species $s$ at plot $p$ in year $t$ for MCMC draw $g$. A temporal trend was then estimated for each species and posterior draw as the ordinary least-squares slope of $\bar{\mu}_{s,t,K}^{(g)}$ against centered year:

$$
\gamma_{s,K}^{(g)} =
\frac{\sum_t (t - \bar{t}) \bar{\mu}_{s,t,K}^{(g)}}{\sum_t (t - \bar{t})^2}
$$

Detection was defined as the posterior probability that the absolute slope exceeded a biologically meaningful threshold:

$$
P\left(|\gamma_{s,K}^{(g)}| \ge \frac{\delta}{T}\right) \ge 0.80
$$

where $\gamma_{s,K}^{(g)}$ is the estimated trend slope for species $s$ at sample size $K$, $\delta = 0.20$ corresponds to a cumulative 20\% change over the window, and $T = t_{\max} - t_{\min}$ of the window, defined as the final year minus the first year. Plots were sampled independently within each year of the window, allowing different plots to contribute to the trend estimate in different years.

Observed trend detectability answers the question: given the trends already present at a site, how many plots would be needed to detect them with high probability?


### Imposed trend detection

The imposed trend detectability analysis evaluates the ability to detect a trend of known magnitude that may not be present in the observed data. This provides a prospective analysis analogous to the counterfactual change detection analysis, but for gradual directional trends rather than discrete changes at a single time point. For each window, an additive linear trend was imposed on the posterior predictive means before estimating slopes:

$$
\mu_{s,p,t}^{\text{trend},(g)} =
\mu_{s,p,t}^{(g)} + \frac{\delta}{T}(t - t_0)
$$

where $\mu_{s,p,t}^{(g)}$ is the original posterior predictive cover, $\mu_{s,p,t}^{\text{trend},(g)}$ is the prediction after imposing the trend, $t_0$ is the first year of the window, $\delta$ is the imposed cumulative change, and $T$ is the calendar time span of the window. Slopes $\gamma_{s,K}^{(g)}$ were then estimated from the trend-modified predictions using the same centered-year regression described above. Detection was assessed using the same posterior probability criterion as in the observed trend analysis.

Imposed trend detectability answers the question: if a trend of the specified magnitude were to emerge at a site, how many plots would be required to detect it?


### Slope estimation stability

The detection-based analyses above evaluate whether an estimated slope exceeds a fixed threshold. A complementary question is whether the slope estimate itself is stable—that is, whether it would yield a similar value if a different set of plots were selected. This is directly relevant to monitoring design: a slope estimate that varies substantially depending on which plots are sampled provides a weak basis for inference, regardless of whether it exceeds a detection threshold.

To evaluate this, we computed the coefficient of variation (CV) of the slope estimate across Monte Carlo replicates at each sample size $K$. For each species $s$, window position, and sample size, each of the $R = 50$ Monte Carlo replicates yields a mean slope (averaged across posterior draws):

$$
\bar{\gamma}_{s,K,m} = \frac{1}{G} \sum_{g=1}^{G} \gamma_{s,K,m}^{(g)}
$$

where $\gamma_{s,K,m}^{(g)}$ is the slope estimate for species $s$ at sample size $K$ in Monte Carlo replicate $m$ for posterior draw $g$, and $G$ is the number of posterior draws.

The coefficient of variation across replicates is then:

$$
\text{CV}_s(K) = 
\frac{\text{SD}\left(\bar{\gamma}_{s,K,1}, \ldots, \bar{\gamma}_{s,K,R}\right)}
{\left|\text{mean}\left(\bar{\gamma}_{s,K,1}, \ldots, \bar{\gamma}_{s,K,R}\right)\right|}
$$

At low $K$, slope estimates vary widely depending on which plots are selected (high CV); as $K$ increases, estimates converge (low CV). A CV below 0.50 indicates that the standard deviation of the slope estimate is less than half its magnitude, such that variation in plot selection does not qualitatively change the inferred direction and strength of the trend.

This stability analysis provides a perspective complementary to the detection-based analyses. Adding spatially heterogeneous plots to annual site means can increase year-to-year variability, making threshold-based detection more difficult at higher $K$—an effect that reflects real ecological heterogeneity but complicates interpretation of detection curves. In contrast, the CV metric evaluates the precision of the slope estimate itself and decreases monotonically with $K$ at nearly all sites.

For each species and site, we defined $K_{\text{CV50}}$ as the smallest sample size at which $\text{CV}_s(K) \le 0.50$, representing the point at which slope estimates become stable with respect to plot selection.


### Definition of $N^*$ for trend detection

For each site, species, and window width, we defined $N^*$ as the smallest number of plots $K$ for which the trend detection criterion was met. As with the counterfactual change detection analysis, $N^*$ values were computed separately for each species and then summarized at the community level.

<!-- FIGURE PLACEHOLDER -->
<!-- Example trend detection power curves: slope detectability vs K for selected species at an example site -->


## Scope and relationship between analyses

The counterfactual year-to-year change detection and trend analyses address ecological change at different temporal scales. The counterfactual year-to-year change detection analysis quantifies the spatial replication required to detect a specified relative change at a single time point, providing guidance relevant to monitoring for abrupt ecological shifts. The trend analysis evaluates the spatial replication required for reliable trend estimation over multi-year windows, providing guidance relevant to tracking gradual directional changes.

These analyses are complementary. The slope stability analysis from the trend framework directly parallels the counterfactual change detection curves: both ask how the quality of inference improves with additional plots. Where both analyses converge on similar sample size requirements, the recommendation is strengthened.


# RESULTS

We present results from site-level analyses across NEON, drawing on both the counterfactual change detection framework and the multi-year trend analysis. Together, these analyses evaluate how the ability to detect ecological change and estimate directional trends depends on spatial replication (number of plots) under realistic levels of spatial and temporal variability.

The analyses leverage 8–12 years of NEON vegetation data per site, representing a substantial increase in temporal replication relative to the 2018 design optimization effort. This expanded record enables direct evaluation of both year-to-year change detection and multi-year trend estimation within a unified modeling framework.

We begin with a summary of site-level data characteristics, including the number of years of observation and the number of species retained after applying the cover filter. We then present detailed results for a small number of representative sites to illustrate the behavior of each analysis, followed by summary tables that synthesize results across all completed sites.


### Site characteristics

Across completed sites, the number of years of observation ranged from 2013 to 2024, with most sites having at least eight years of data. The number of species included in the GJAM model ranged from 39 to 406 per site, reflecting substantial variation in species richness across the Observatory. After applying the 3% mean cover threshold, the number of species retained for analysis was reduced to 4 to 122 per site.

All species were included in model fitting and posterior prediction to preserve community structure. Detection and stability metrics, however, were restricted to species with mean raw cover ≥ 3%. Below this threshold, a 20% relative change falls within the resolution of the field protocol and cannot be reliably distinguished from measurement noise. For example, a species with 2% mean cover would require detection of a 0.4 percentage point change under a 20% perturbation, which is smaller than the effective resolution of visual percent cover estimates in the field. This filtering ensures that inference is focused on the portion of the community where meaningful change can be resolved, and the same filtered species set was used consistently across both the change detection and trend analyses.

For each site, we report three species counts: (1) the total number of species observed in the raw dataset, (2) the number of species included in the model, and (3) the number of species retained for inference after applying the 3% mean cover threshold (Table 1).

```{r site-summary-table, echo=FALSE, message=FALSE}

site_summary <- read_csv("siteSummaryIntroDiscssion.csv")

kable(
  site_summary,
  caption = "Summary of site characteristics, including years of observation and species counts for observed, modeled, and inference subsets.",
  booktabs = TRUE,
  align = "l"
)

```

## Year-to-year change detection (counterfactual analysis)

### Example sites

We illustrate the behavior of the counterfactual change detection analysis using three representative sites: ABBY, GRSM, and BONA (Figure 2), spanning a range of community composition and spatial heterogeneity.

At all three sites, detection increased with sample size at the focal effect size of $\delta = 0.20$, but the rate of increase differed among sites. At ABBY, detection rose steadily with increasing $K$, approaching high detection probabilities at larger sample sizes. GRSM required a greater number of plots to achieve comparable levels of detection, reflecting increased spatial variability. At BONA, detection increased with $K$ but did not reach the 80% threshold within the evaluated range, indicating that the imposed change could not be reliably detected even at maximum sampling effort.

At smaller effect sizes ($\delta = 0.15$), detection remained low across all sample sizes, indicating that changes below 20% are not reliably detectable under observed spatial variability. At larger effect sizes ($\delta \ge 0.25$), detection increased rapidly and approached saturation at relatively small sample sizes, particularly at ABBY.

```{r fig-power-curves, echo=FALSE, results='asis'}
cat("
\\begin{figure}[!htbp]
\\centering
\\includegraphics[width=0.82\\textwidth]{fig_power_curves_ABBY_GRSM_BONA.png}
\\caption{Counterfactual change detection across three example sites. Curves show the percentage of species with detection probability at least 0.80 as a function of the number of plots sampled per year ($K$), for imposed relative changes of 15--30\\%.}
\\label{fig:power_curves}
\\end{figure}
")
```

### Summary across sites

Across sites, the number of plots required to achieve the detection criterion ($K^*$) varied substantially (Table 2). At $\delta = 0.20$, most sites reached the 80\% detection threshold between approximately $K = 20$ and $K = 30$, although several sites did not reach the threshold within the evaluated range.

Detection increased monotonically with sample size at nearly all sites, but the rate of increase differed. Sites with relatively homogeneous spatial structure showed rapid gains in detection with increasing $K$, whereas more heterogeneous communities exhibited slower, more gradual increases.

Detection at the maximum evaluated sample size ($K = 30$) also varied among sites, indicating that for some communities the 80\% detection threshold was not achievable within the range of sampling effort considered. These differences reflect variation in community composition and spatial heterogeneity rather than differences in model behavior.

Species-level detection requirements were substantially lower than the community-level thresholds reported here, indicating that many species are individually detectable at small sample sizes. However, the community-level metric requires that most species exceed the detection criterion, resulting in larger sample size requirements for reliable inference at the community scale.

```{r site-summary-table tier0, echo=FALSE, message=FALSE}

tier0_summary_tbl <- read_csv("tier0_summary_pub.csv")

kable(
  tier0_summary_tbl,
  caption = "Summary of counterfactual change detection across sites. K* is the smallest evaluated number of plots at which at least 80% of species exceeded the detection criterion for a 20% relative change. Detection at K = 30 reports the percentage of species detectable at the maximum evaluated sample size.",
  booktabs = TRUE,
  align = "l"
)

```


## Multi-year trend detection

### Example sites

Patterns of trend detectability broadly paralleled those observed for year-to-year change detection, but with important differences arising from the temporal structure of the signal (Figure 3).

At ABBY, trend detection increased consistently with sample size, and both detection and stability were achieved at relatively small $K$. In contrast, GRSM required larger sample sizes to achieve reliable detection of trends, reflecting increased variability in annual site means. BONA exhibited relatively strong trends but required larger sample sizes to stabilize slope estimates, indicating that strong signals do not necessarily translate to stable inference.

```{r fig-trend-detection, echo=FALSE, results='asis'}
cat("
\\begin{figure}[!htbp]
\\centering
\\includegraphics[width=0.82\\textwidth]{fig_trend_detection_ABBY_GRSM_BONA.png}
\\caption{Trend detection across three example sites. Curves show the fraction of species with detection probability at least 0.80 as a function of the number of plots sampled per year ($K$), for imposed cumulative trends of different magnitudes.}
\\label{fig:trend_detection}
\\end{figure}
")

```

Add some stability text

```{r fig-trend-stability, echo=FALSE, results='asis'}
cat("
\\begin{figure}[!htbp]
\\centering
\\includegraphics[width=0.82\\textwidth]{fig_trend_stability_CV_ABBY_GRSM_BONA.png}
\\caption{Slope estimation stability across three example sites. Curves show the median coefficient of variation (CV) of species-level slope estimates as a function of the number of plots sampled per year ($K$). Lower CV indicates greater stability with respect to plot selection. Horizontal lines indicate reference thresholds for moderate (CV = 0.50) and high (CV = 0.25) stability.}
\\label{fig:trend_stability}
\\end{figure}
")
```

### Detection and stability

Detection and stability provided complementary perspectives on trend inference. Detection evaluates whether the estimated slope exceeds a specified threshold, while stability evaluates whether the slope estimate is consistent across repeated subsamples of plots.

Across sites, detection thresholds were typically reached at moderate sample sizes, but stability required larger values of $K$. In many cases, slope estimates remained sensitive to plot selection even after detection thresholds were met.

The coefficient of variation of slope estimates decreased monotonically with increasing $K$ at nearly all sites (Figure 4), indicating that additional plots consistently improve the precision of trend estimates. This contrasts with detection-based metrics, which can be influenced by spatial heterogeneity in ways that complicate interpretation.

### Summary across sites

Across the Obseratory, sample size requirements for trend detection and stability varied substantially among sites (Table 3). Sites with greater species richness and spatial heterogeneity generally required more plots to achieve stable and detectable trends.

In some cases, detection and stability metrics converged on similar sample size requirements, providing strong guidance for monitoring design. In others, detection thresholds were reached at smaller sample sizes than stability thresholds, indicating that reliable estimation of trend magnitude requires more intensive sampling than detection alone.

Across sites, the number of plots required to detect a 20% cumulative trend (Trend K*) and the number required to achieve stable slope estimates (K_CV50) differed substantially (Table X). In most cases, detection thresholds were reached at smaller sample sizes than stability thresholds, indicating that while directional change can often be detected with moderate sampling effort, reliable estimation of trend magnitude requires additional plots.

At some sites, the detection threshold was not achieved within the evaluated range of sample sizes, even though slope estimates became progressively more stable with increasing K. Conversely, sites with relatively low spatial heterogeneity showed similar sample size requirements for detection and stability. Together, these results indicate that detectability and stability represent distinct constraints on inference, and both must be considered when evaluating monitoring design.

```{r site-summary-table trendrun, echo=FALSE, message=FALSE}

trendrun_summary_tbl <- read_csv("trendrun_summary_results_table.csv")

kable(
  tier0_summary_tbl,
  caption = "Summary of multi-year trend detection and slope stability across sites. Trend K* is the smallest evaluated number of plots at which at least 80% of species exceeded the detection criterion for a 20% imposed cumulative trend. Detection at K = 30 reports the percentage of species detectable at the maximum evaluated sample size. K_CV50 is the smallest evaluated number of plots at which the median slope coefficient of variation was less than or equal to 0.50; CV at K = 30 reports the median slope CV at the maximum evaluated sample size.",
  booktabs = TRUE,
  align = "l"
)

```


## Synthesis across analyses

The counterfactual change detection and trend analyses provide complementary perspectives on the detectability of ecological change. Both analyses show that increasing the number of sampled plots improves inference, but the magnitude of this improvement depends strongly on community structure and spatial heterogeneity.

Where both approaches converge on similar sample size requirements, confidence in those estimates is strengthened. Where they diverge, the difference reflects the distinct challenges of detecting discrete changes versus estimating continuous trends. Together, these results provide a framework for evaluating how sampling design constrains inference across multiple temporal scales.




[1–2 sites only — figures]

- power curves
- maybe species heatmap (optional)

### Summary across sites

[table outputs from tierp0 script]

- power_table_all_sites.csv
- mde_equiv_table
- nstar_table


## Multi-year trend detection and stability

Where the counterfactual analysis evaluates the detectability of discrete changes, the trend analysis addresses a complementary question: whether directional change in species abundance can be reliably estimated over time. In this context, the focus shifts from detecting a difference between two states to estimating the rate of change across multiple years.

As with the counterfactual analysis, inference is based on posterior predictive draws from a single site-level model. For each species, predicted cover values are generated for all plot–year combinations, preserving the spatial and temporal structure of the observed data. These predictions are then used to estimate annual site-level means under varying levels of spatial replication.

To evaluate the influence of sampling intensity, plots are subsampled independently within each year, and site-level mean cover is computed for each posterior draw. A linear trend is then estimated across years for each species, yielding a posterior distribution of slope parameters that reflects both model uncertainty and the variability introduced by spatial subsampling.

Two complementary properties of the estimated trend are evaluated. First, detection assesses whether the magnitude of the estimated slope exceeds a specified threshold, analogous to the threshold-based detection used in the counterfactual analysis. Second, stability evaluates the sensitivity of the estimated slope to the selection of plots, quantified as the coefficient of variation of slope estimates across repeated subsamples at each sample size.

Together, these metrics describe both the ability to detect directional change and the reliability of the estimated rate of change. As in the counterfactual analysis, the objective is not to optimize model fit, but to quantify how spatial sampling constrains inference about ecological change.






## Multi-year trend analysis

### Slope estimation stability (primary result)

### Example sites

- stability curves (CV vs K)
- % stable species

### Summary across sites

- stability_cv_table.csv
- K_cv50 (you already compute this)

---

### Trend detectability (supplementary)

(short, clearly secondary)

---

### Plot continuity and sampling variation

(optional but useful — you already built it)

- plot_continuity_table.csv





### Community-level power curves

<!-- FIGURE PLACEHOLDER -->
<!-- fig:power_curves — Power curves (pct species detectable vs K) for HARV, OSBS, KONZ, GRSM at effect sizes 15-30% -->

At all example sites, community-level detection power increased monotonically with sample size at the focal effect size of $\delta = 0.20$ (20\% relative change). After applying the 3\% cover floor, the number of species included in the community metric ranged from 45 (HARV) to 55 (OSBS).

<!-- TABLE PLACEHOLDER -->
<!-- tab:power_table — full power table from power_table_all_sites.csv -->

At $\delta = 0.20$, all four sites crossed the 80\% community detection threshold between $K = 20$ and $K = 25$ plots: HARV at $K = 25$ (80\%), OSBS at $K = 20$ (83\%), KONZ at $K = 25$ (80\%), and GRSM at $K = 20$ (82\%). At $\delta = 0.15$ (15\% relative change), detection remained below 45\% even at $K = 30$, indicating that changes below 20\% are not reliably detectable at any sample size under the current framework. At $\delta = 0.25$, detection was nearly saturated ($>$90\%) at all sample sizes including $K = 5$.

### Year-to-year stability

Detection power was stable across baseline years within each site. At $\delta = 0.20$ and $K = 20$, the standard deviation of community detection across years was 0--2.4 percentage points (KONZ: 0\%, HARV: 1.8\%, GRSM: 2.2\%, OSBS: 2.4\%). This stability confirms that the paired counterfactual design successfully eliminates year-to-year confounds.


\subsection{Multi-year trend analysis}

### Slope estimation stability (primary result)

<!-- FIGURE PLACEHOLDER -->
<!-- fig:stability — Slope CV vs K for example sites, showing monotonic decrease -->

The slope stability analysis showed that trend estimates became progressively more reliable as sample size increased. Across the network, the median coefficient of variation (CV) of species-level slope estimates decreased monotonically with $K$ at virtually all sites for both 5-year and 10-year windows. The number of plots required for stable trend estimation (median CV $< 0.50$) ranged from 5 at spatially homogeneous sites (e.g., arid grasslands) to 20 or more at topographically complex sites (e.g., montane forests), providing genuine site-level differentiation.

<!-- TABLE PLACEHOLDER -->
<!-- tab:stability — Stability CV table from stability_cv_table.csv -->

These results complement the perturbation-based findings: both analyses identify $K = 20$--25 as the sample size at which inference quality transitions from unreliable to adequate at most sites, albeit through different metrics (detection power vs.\ estimation precision).


### Imposed trend detectability (supplementary)

<!-- FIGURE PLACEHOLDER -->
<!-- fig:sensB — Community detection by K and imposed trend at example sites -->

The detection-based trend analyses revealed a spatial dilution effect at a number of sites: detection probability decreased rather than increased with additional plots. This occurs because adding spatially heterogeneous plots increases the year-to-year scatter of annual site means, making OLS slopes harder to distinguish from zero against a fixed threshold. This effect is ecologically real---more plots sample more habitat variation---and represents a fundamental difference between trend detection and point-in-time change detection. Detection-based trend power curves should therefore be interpreted with this caveat.

### Plot continuity and the COVID-2020 gap

<!-- TABLE PLACEHOLDER -->
<!-- tab:plot_continuity — Plot continuity table from plot_continuity_table.csv -->

The COVID-2020 sampling disruption reduced the number of plots sampled at many sites, creating continuity gaps that affect multi-year trend analysis. To address this, the trend analysis samples plots independently for each year within a window, rather than requiring the same plots to be sampled in every year. This design reflects realistic monitoring conditions where plot availability may vary across years.


\section{\uppercase{Discussion}}

add something about impact of filtering

add text about trend run inversion in vs K

add text about stability threshold and how to interpret; CV threshold ↔ N* conceptually (very powerful in Discussion)


### Complementary perspectives on sampling adequacy

This analysis demonstrates that conclusions about required spatial replication depend on both the magnitude of ecological change and the temporal scale over which it is assessed. The perturbation-based power analysis quantifies the ability to detect a specified relative change in species cover at a single time point, providing guidance relevant to monitoring for abrupt ecological shifts---invasive species establishment, disturbance events, or rapid phenological changes. The trend stability analysis quantifies the number of plots required for reliable estimation of multi-year directional trends, providing guidance relevant to tracking gradual changes in community composition driven by climate, land use, or successional dynamics.

Both analyses converge on a consistent recommendation: 20--25 plots are required for reliable ecological inference. The perturbation analysis identifies this range as the sample size at which 80\% of species (above the 3\% cover floor) are detectable at a 20\% relative change. The trend stability analysis identifies this range as the sample size at which slope estimates become stable (CV $< 0.50$ for most species). The convergence of these independent analyses---addressing different temporal scales through different statistical frameworks---strengthens confidence in the recommendation.


### Per-species relative framework

The use of per-species perturbation and detection thresholds scaled to each species' raw mean cover represents an important methodological choice. By imposing a 20\% relative change on every species (rather than a fixed absolute change), the analysis ensures that low-cover and high-cover species face equivalent proportional challenges. A species at 10\% mean cover receives a 2 percentage point perturbation against a 1 percentage point threshold; a species at 60\% cover receives 12 percentage points against 6. This scaling avoids the bias inherent in fixed absolute frameworks, where a given percentage-point change is ecologically trivial for dominant species but dramatic for subordinate ones.

The 3\% cover floor reflects the practical limit of both field measurement precision (NEON records cover to the nearest 1\%) and model prediction reliability at low cover. Below this threshold, a 20\% relative change falls within the recording resolution of the field protocol. At most NEON sites, the majority of species fall below the 3\% floor, meaning that the power and stability analyses characterize detection capability for the more abundant portion of the community.


### Modeling framework and dimension reduction

The use of GJAM as the underlying modeling framework enables joint inference across all species at a site, accounting for shared environmental responses and residual covariance among species. This represents a substantial methodological advance over the univariate mixed-effects models employed in the 2018 optimization analysis (Barnett et al. 2018), which treated each species independently and could not account for the coordinated responses that characterize plant communities.

The dimension reduction framework (Taylor-Rodriguez et al. 2017) was essential for applying this approach across the full diversity gradient of NEON sites. Without reduction, the $S \times S$ covariance matrix cannot be estimated reliably when the number of species exceeds the effective sample size, a condition met at the majority of NEON sites. By representing the covariance through a reduced factor structure, the model captures the dominant modes of inter-species association while remaining computationally tractable at sites with hundreds of species.


### Spatial dilution in trend detection

The detection-based trend analyses revealed an important finding: at many sites, the probability of detecting a trend against a fixed threshold decreased as more plots were added. This spatial dilution effect occurs because additional plots sample more habitat variation, increasing the year-to-year scatter of annual site means and making OLS slopes harder to distinguish from zero. This effect is ecologically real and represents a fundamental difference between trend detection against a fixed threshold and point-in-time change detection, where averaging across more plots always reduces noise.

The slope stability analysis was developed to complement the detection-based approach. Because stability measures the precision of the slope estimate (via the coefficient of variation across plot resamples) rather than its exceedance of a threshold, it does not suffer from the spatial dilution artifact. More plots always provide a more representative sample of the site, reducing the sensitivity of the trend estimate to which specific plots were selected.


### Implications for NEON sampling design

The results indicate that NEON's current design of 20--30 plots per site is well-calibrated for detecting 20\% relative changes in plant species cover and for producing stable trend estimates. At four validation sites spanning temperate deciduous forest, tallgrass prairie, montane forest, and subtropical pine flatwoods, all sites crossed the 80\% community detection threshold at $K = 20$--25 for a 20\% relative effect. Across the full network, most sites showed consistent patterns, with site-level variation in trend stability providing genuine differentiation: spatially homogeneous sites (arid grasslands, agricultural systems) achieve stable trend estimates with as few as 5 plots, while topographically complex sites require 15--20 or more.

Changes below 15\% relative are not reliably detectable at any sample size under the current framework, establishing a practical floor on the ecological sensitivity of NEON's plot-based monitoring. Changes of 25\% or larger are detectable with as few as 5 plots, indicating that major ecological events will be captured regardless of sampling intensity.


### Addressing the 2018 trend detection question

The 2018 design optimization effort identified trend detection as a priority question but determined that it could not be addressed with the limited temporal replication then available (Barnett et al. 2018). With 8--12 years of data now available at most sites, the moving-window trend analysis presented here directly addresses that outstanding question. The slope stability framework provides actionable guidance: 20--25 plots are sufficient for reliable trend estimation over 5--10 year windows at most sites, consistent with the perturbation-based change detection results.


### Limitations and future directions

Several limitations of the current analysis warrant emphasis. First, the model is fit once per site using all available data, and posterior predictions are generated conditional on this single fit (single-fit design). The analysis does not propagate uncertainty arising from plot selection into the model fitting step, meaning that power estimates represent upper bounds on detection capability. The consistency of recommendations across ecologically distinct sites and with the 2018 univariate analysis provides confidence that the directional findings---20--25 plots for reliable change detection and trend estimation---are robust, though the true sample size requirement may be somewhat higher. A limited refit analysis, in which the model is re-estimated at each sample size, could quantify this optimism and is identified as a priority for future work.

Second, multiple relative effect sizes (15--30\%) were tested, revealing that 20\% is the critical threshold where sample size determines detection capability. Sensitivity to alternative detection threshold values (currently 10\% of raw cover) was not systematically evaluated but could be addressed in future work.

Third, the GJAM-based approach treats observed cover values as compositional, normalizing species shares within each observation. This compositional normalization compresses absolute cover values relative to raw field data, which is why the cover floor and per-species scaling are computed from raw field observations rather than model predictions.

The analytical framework developed here is extensible in several directions. Longer NEON time series will enable evaluation of trend detection over wider windows and provide empirical validation of the power estimates. The GJAM-based approach could also be extended to other NEON observation systems where multivariate community data are collected, including small mammals, ground beetles, and soil microorganisms.



\section{\uppercase{References}}
Claire Lunch, Christine Laney, Nathan Mietkiewicz, Eric Sokol, Kaelin Cawley and NEON (National Ecological Observatory Network) (2024). neonUtilities: Utilities for Working with NEON Data. R package version 2.4.3. https://doi.org/10.32614/CRAN.package.neonUtilities

Barnett, D. T., P. A. Duffy, D. S. Schimel, R. E. Krauss, E. I. Azuaje, K. M. Irvine, F. W. Davis, A. E. Gelfand, A. S. Thorpe, D. Gudex-Cross, J. M.   \hangindent= 1cm
McKay, J. T. McCorkel and C. L. Meier. The terrestrial and organism and biogeochemistry spatial sampling design for the National Ecological             Observatory Network. Ecosphere. 

Bolker, M. B., M. E. Brooks, C. J. Clark, S. W. Geange, J. R. Poulsen, M. H. H. Stevens, J. S. White. 2008. Generalized linear mixed models: a practical \hangindent= 1cm
guide for ecology and evolution. Trends in Ecology and Evolution 24(3):127-136. 

Chao, A., Gotelli, N.J., et al., 2014. Rarefaction and extrapolation with Hill numbers: A framework for sampling and estimation in species diversity \hangindent= 1cm
studies. Ecological Monographs, 84(1), pp.45-67.

Chao, A. et al., 2015. Unveiling the species-rank abundance distribution by generalizing the Good-Turing sample coverage theory. Ecology, 96(5), \hangindent= 1cm
pp.1189-1201.

Clark, J, S., D. Nemergut, B. Seyednasrollah, P. J. Turner, and S. Zhang. 2017. Generalized joint attribute modeling for biodiversity analysis: median-zero, multivariate, multifarious data. Ecological Monographs \hangindent= 1cm
87(1):34-56.

Colwell, R.K. & Coddington, J.A., 1994. Estimating Terrestrial Biodiversity through Extrapolation. Philosophical Transactions of the Royal Society of \hangindent= 1cm
London. Series B: Biological Sciences, 345(1311), pp.101-118. 

Colwell, R.K. et al., 2012. Models and estimators linking individual-based and sample-based rarefaction, extrapolation and comparison of assemblages. \hangindent= 1cm
Journal of Plant Ecology, 5(1), pp.3-21.

Crowley, P. H. 1992. Resampling methods for computation-intensive data analysis in ecology and evolution. Annual Review of Systematics, 23:405-447.
\hangindent= 1cm

Elmendorf, S. C., K. D. Jones, B.I . Cook, J. M. Diez, C. A. F. Enquist, M.O. Jones, R. A. Hufft, S. J. Mazer, A. J. Miller-Rushing, D. Moore, M. D. \hangindent= 1cm
Schwartz, and J. F. Weltzin. 2016. The Plant Phenology sampling design for the National Ecological Observatory Network. Ecosphere. 7(4):e01303. 10.1002/ecs2.1303.

Gotelli, N.J. & Colwell, R.K., 2001. Quantifyinf biodiversity: procedures and pitfalls in the measurement and comparison of species richness. Ecology \hangindent= 1cm
Letters, 4 Abstract(4), pp.379-391. 

Gotelli, Nicholas J.; Colwell, R.K., 2011. Estimating species richness. In B. J. Magurran, Anne E.; McGill, ed. Biological diversity frontiers in \hangindent= 1cm
measurement and assessment. Oxford University Press, pp. 39-54. 

Hinckley, E. S., G. B. Bonan, G. J. Bowen, B. P. Colman, P. A. Duffy, C. L. Goodale, B. Z. Houlton, E. Marin-Spiotta, K. Ogle, S. V. Ollinger, E. A. \hangindent= 1cm
Paul, P. M. Vitousek, K. C. Weathers, and D. G. Williams. 2016. The soil and plant biogeochemistry sampling design for The National Ecological Observatory Network Ecosphere. 7(3):e01234.

Hsieh, T.C., Ma, K.H. & Chao, A., 2016. iNEXT: an R package for rarefaction and extrapolation of species diversity (Hill numbers). Methods in Ecology \hangindent= 1cm
and Evolution, 7, pp.1451-1456.

Kain, M. P., Bolker, B. M., M. W. McCoy. 2015. A practical guide and power analysis for GLMMs: detecting among treatment variation in random effects. \hangindent= 1cm
PeerJ 3:e1226; DOI 10.7717/peerj.1226.

Kao, R. H., C. M. Gibson, R. E. Gallery, C. L. Meier, D. T. Barnett, K. M. Docherty, K. K. Blevins, P. D. Travers, A. Azuaje, Y. P. Springer, K. M. \hangindent= 1cm
Thibault, V. J. McKenzie, M. Keller, L. F. Alves, E. S. Hinckley, J. Parnell, and D. Schimel. 2012. NEON terrestrial field observations: designing continental-scale, standardized sampling. Ecosphere 3(12):1-17.

Keller, M., D. S. Schimel, W. W. Hargrove, and F. M. Hoffman. 2008. A continental strategy for the national ecological observatory network. Frontiers in \hangindent= 1cmEcology and Environment 6:282-284.

Meier, C. L, K. M. Thibault, and D. T. Barnett. 2023. Spatial and temporal sampling strategy connecting NEON Terrestrial Observation System protocols. \hangindent= 1cm
Ecosphere 14(3):e4455. https://doi.org/10.1002/ecs2.4455

Peet, R. K., T. R. Wentworth, and P. S. White. 1998. A flexible, multipurpose method for recording vegetation composition and structure. Castanea \hangindent= 1cm63(3):262-274.

Ogle, K., and J. F. Reynolds. 2004. Plant responses to precipitation in desert ecosystems: integrating functional types, pulses, thresholds, and delays. \hangindent= 1cm
Oecologia 141:282:294.

Schimel, D., M. Keller, S. Berukoff, R. Kao, H. Loescher, H. Powell, T. Kampe, D. Moore, W. Gram, D. Barnett, R. Gallery, C. Gibson, K. Goodman, C. \hangindent= 1cm
Meier, S. Parker, L. Pitelka, Y. Springer, K. Thibault, and R. Utz. 2011. 2011 science strategy: Enabling continental-scale ecological forecasting.

Shmida, A. 1984. Whittaker's plant diversity sampling method. Israel Journal of Botany 33:41-46.

Stohlgren, T. J., M. B. Falkner, and L. D. Schell. 1995. A modified-Whittaker nested vegetation sampling method. Vegetatio 117:113-121.
\hangindent= 1cm

Taylor-Rodriquez, D., K. Kaufeld, E. M. Schliep, J. S. Clark, and A. E. Gelfand. 2017. Joint Species Distribution Modeling: Dimension reduction using dirichlet processes. Bayesian Analysis. 12(4):939:967.
\hangindent= 1cm

Thompson, S. K. 2012. Sampling. Third edition. Wiley, New Jersey, USA.

Thorpe, A. S., D. T. Barnett, S. C. Elmendorf, E.-L. S. Hinckley, D. Hoekman, K. D. Jones, K. E. LeVan, C. L. Meier, L. F. Stanish, and K. M. Thibault. \hangindent= 1cm
2016. Introduction to the sampling designs of the National Ecological Observatory Network Terrestrial Observation System. Ecosphere 7(12):e01627. 10.1002/ecs2.1627

USDA, NRCS. 2016. The PLANTS Database (http://plants.usda.gov, 23 January 2016). National Plant Data Team, Greensboro, NC 27401-4901 US.

Zuur A.F., Ieno E.N., Walker Saveliev A.A., Smith N.J., G.M. (2009) Mixed effects models and extensions in ecology with R. Statistics for Biology and \hangindent= 1cm
Health. Springer, New York, NY.

Clark, J. S., D. Nemergut, B. Seyednasrollah, P. J. Turner, and S. Zhang. 2017. Generalized joint attribute modeling for biodiversity analysis: \hangindent= 1cm
median-zero, multivariate, multifarious data. Ecological Monographs 87:34--56.

Taylor-Rodriguez, D., K. Kaufeld, E. M. Schliep, J. S. Clark, and A. E. Gelfand. 2017. Joint species distribution modeling: dimension reduction using \hangindent= 1cm
Dirichlet processes. Bayesian Analysis 12:939--967.

Urquhart, N. S., S. G. Paulsen, and D. P. Larsen. 1998. Monitoring for policy-relevant regional trends over time. Ecological Applications 8:246--257.
\hangindent= 1cm

NEON (National Ecological Observatory Network). Plant presence and percent cover (DP1.10058.001), RELEASE-2025. https://doi.org/10.48443/zp67-s364. Dataset \hangindent= 1cm
accessed from https://data.neonscience.org/data-products/DP1.10058.001/RELEASE-2025 on January 3, 2026.
