---
title: "Detecting Year-to-Year Change in Plant Cover: Baseline and Sample-Size Sensitivity Using NEON Data"
author: "Dave Barnett, Courtney Meier, and Eric Sokole"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_caption: true
  rmarkdown::pdf_document:
    number_sections: true
params:
  example_site: "JERC"
  top_n_species: 6
  min_sample_size: 5   # floor for plots shown on x-axes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 8, fig.height = 5
)

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2)
  library(stringr); library(knitr); library(purrr); library(arrow)
})

# --- simple, sequential figure counter so captions are 1,2,3,... (not 3.1, 3.2) ---
.fig_counter <- 0L
figcap <- function(txt) {
  .fig_counter <<- .fig_counter + 1L
  paste0("Figure ", .fig_counter, ". ", txt)
}

# --- helper: harmonize schemas new/old (cwm_mean vs weighted_detection, year_pair, etc.) ---
sanitize_community <- function(df) {
  if (!is.data.frame(df) || !nrow(df)) return(df)
  if (!"weighted_detection" %in% names(df) && "cwm_mean" %in% names(df)) {
    df <- dplyr::rename(df, weighted_detection = cwm_mean)
  }
  if (!"year_pair" %in% names(df) &&
      all(c("year_baseline","year_changed") %in% names(df))) {
    df <- dplyr::mutate(df, year_pair = paste0(.data$year_baseline, "_", .data$year_changed))
  }
  df
}

# --- helper: replicate-averaged curves (mirrors your helpers) ---
rep_avg_fn <- function(df) {
  stopifnot(all(c("site","sample_size","replicate","weighted_detection") %in% names(df)))
  if (!"year_pair" %in% names(df)) {
    stop("Expected `year_pair`.", call. = FALSE)
  }
  has_ci <- all(c("ci_lower","ci_upper") %in% names(df))
  df %>%
    mutate(
      ci_lower = if (has_ci) as.numeric(ci_lower) else NA_real_,
      ci_upper = if (has_ci) as.numeric(ci_upper) else NA_real_
    ) %>%
    group_by(site, sample_size, year_pair) %>%
    summarise(
      mean_detection = mean(weighted_detection, na.rm = TRUE),
      n_replicates   = dplyr::n_distinct(replicate),
      ci_lower_mean  = if (has_ci) mean(ci_lower, na.rm = TRUE) else NA_real_,
      ci_upper_mean  = if (has_ci) mean(ci_upper, na.rm = TRUE) else NA_real_,
      .groups = "drop"
    )
}

# --- helper: summarise across year pairs (site × N) ---
site_avg_fn <- function(rep_avg_df) {
  rep_avg_df %>%
    group_by(site, sample_size) %>%
    summarise(
      detection_mean = mean(mean_detection, na.rm = TRUE),
      detection_sd   = sd(mean_detection, na.rm = TRUE),
      ci_lower       = quantile(mean_detection, 0.025, na.rm = TRUE),
      ci_upper       = quantile(mean_detection, 0.975, na.rm = TRUE),
      n_year_pairs   = dplyr::n_distinct(year_pair),
      .groups = "drop"
    )
}

# --- helper: fraction of year pairs passing threshold (site × N) ---
fraction_passing <- function(rep_avg_df, threshold = 0.8) {
  rep_avg_df %>%
    mutate(pass = mean_detection >= threshold) %>%
    group_by(site, sample_size) %>%
    summarise(
      prop_yearpairs_passing = mean(pass, na.rm = TRUE),
      mean_at_size           = mean(mean_detection, na.rm = TRUE),
      sd_at_size             = sd(mean_detection, na.rm = TRUE),
      .groups = "drop"
    )
}

# --- guard: required objects injected by tar target render() env ---
need <- function(obj) if (!exists(obj, inherits = FALSE)) stop(sprintf("Missing `%s` in render environment.", obj), call. = FALSE)
need("community_detection")
need("full_detection_summary")
need("relative_cover_df")
has_baseline       <- exists("community_detection_baseline", inherits = FALSE) && nrow(community_detection_baseline) > 0
has_draws_idx_var  <- exists("draws_idx_variable", inherits = FALSE) && nrow(draws_idx_variable) > 0

# --- constants & params ---
threshold <- 0.8
pair_coverage_target <- 0.75
MIN_N <- params$min_sample_size %||% 5

# --- normalize/prepare inputs ---
cd  <- sanitize_community(community_detection)
cdb <- if (has_baseline) sanitize_community(community_detection_baseline) else NULL

# For variable mode, x-axis should reflect *requested* N when present
cd_var <- cd
if ("requested_sample_size" %in% names(cd_var)) {
  cd_var <- cd_var %>% mutate(sample_size = requested_sample_size)
}

# Filter very small N from plots by default
cd_var_filt <- cd_var %>% filter(sample_size >= MIN_N)
cd_base_filt <- if (!is.null(cdb)) cdb %>% filter(sample_size >= MIN_N) else NULL

# Replicate-averaged datasets
rep_avg_var   <- rep_avg_fn(cd_var_filt)
site_avg_var  <- site_avg_fn(rep_avg_var)
rep_avg_base  <- if (!is.null(cd_base_filt)) rep_avg_fn(cd_base_filt) else NULL
site_avg_base <- if (!is.null(rep_avg_base)) site_avg_fn(rep_avg_base) else NULL

# example site selection
available_sites <- sort(unique(cd_var_filt$site))
example_site <- params$example_site
if (!example_site %in% available_sites && length(available_sites)) example_site <- available_sites[1]

rep_avg_var_site   <- rep_avg_var  %>% filter(site == example_site)
site_avg_var_site  <- site_avg_var %>% filter(site == example_site)
rep_avg_base_site  <- if (!is.null(rep_avg_base))  rep_avg_base  %>% filter(site == example_site) else NULL
site_avg_base_site <- if (!is.null(site_avg_base)) site_avg_base %>% filter(site == example_site) else NULL

# tiny baseline overlay helper
add_baseline_layer <- function(p, df_base, yvar = "mean_detection") {
  if (is.null(df_base) || !nrow(df_base)) return(p)
  p +
    geom_point(data = df_base,
               aes(x = sample_size, y = .data[[yvar]]),
               inherit.aes = FALSE, size = 2, shape = 21, stroke = 0.8) +
    geom_line(data = df_base,
              aes(x = sample_size, y = .data[[yvar]], group = 1),
              inherit.aes = FALSE, linetype = "dotted")
}


```

# 1 Introduction

Effective ecological monitoring requires designs that can detect meaningful change against the backdrop of natural variability, observation error, and spatial heterogeneity. The statistical power to detect change is a joint outcome of the ecological processes that generate variation and the sampling design that captures it. Even with well-calibrated field methods, limited replication can mask real change, while overly intensive designs can consume resources without proportionally increasing detection. Balancing sensitivity and efficiency is thus a central challenge in the design and evaluation of long-term ecological networks.

The National Ecological Observatory Network (NEON) plant diversity program addresses this challenge by implementing a standardized, plot-based sampling design across a continental network of sites (Barnett et al., 2019a; Barnett et al., 2019b). Fixed-location plots are revisited annually using consistent methods for estimating species composition and percent cover, enabling both site-level and cross-site comparisons over decadal time frames. This design supports the dual goals of detecting local ecological change and understanding broader patterns in biodiversity and vegetation structure.

With longer time series now available for many sites, it is possible to evaluate the network’s operational sensitivity using the data it produces. Specifically, we ask:

1. **How well are we doing now?** For each site and consecutive year pair, what is the probability of detecting an interannual change in species percent cover, given the plots actually sampled (the **baseline**)?
2. **Can we do as well with fewer plots?** Under a sample-size sensitivity scenario, if we request a target number of plots $N$ but use whatever is actually available for that year pair  
($n = \min(N, n_{\text{available}})$), how close does performance come to the baseline?

We address these questions using the **Generalized Joint Attribute Model** (GJAM; Clark et al., 2017), a Bayesian joint species distribution framework that models the full multivariate response of species abundances while accounting for correlations among species and uncertainty in estimates. GJAM is well-suited for this application because it can generate posterior predictions for alternative sampling or change scenarios without refitting the model, preserving the interspecific covariance structure and full posterior uncertainty from a single run.

Our simulations impose a 20% increase in cover for a focal species and propagate the resulting posterior predictions through both species-level and community-weighted detection metrics. Relative cover provides the weighting for community aggregation, ensuring that results reflect the abundance structure of each site’s plant community. By grounding the analysis in NEON’s sampling protocols and real-world plot availability, our results offer both a current assessment of detection performance and an empirical basis for exploring design tradeoffs under alternative sampling intensities.



# Methods

To address these questions, we combined NEON plant percent cover observations with the joint species modeling framework of GJAM in a site-by-site analysis. For each site, we fit a single model to all available plot–year combinations, using predictors that reflect the temporal and habitat structure of the sampling design. Posterior predictions from these fitted models were then used to evaluate detection probability under observed sampling effort and under controlled reductions in plot number. By simulating change through posterior prediction rather than refitting, we maintained the estimated species–species covariance structure and uncertainty from the original fit, allowing us to attribute differences in detection directly to sampling intensity. The following sections describe the data, model structure, simulation design, and detection metrics in detail.

We analyzed plot-based plant abundance data from the National Ecological Observatory Network (NEON) Plant presence and percent cover product (DP1.10058.001). Data were aggregated to plot × year × species for each site, retaining the original percent cover estimates as the response variable in modeling. NEON’s vegetation protocol (Barnett et al., 2019a) samples species composition and cover in a fixed network of 20 × 20 m plots distributed according to NEON’s terrestrial spatial sampling design (Barnett et al., 2019b). In this study, we used only the Distributed Base plots, which are located across each site to capture landscape-scale environmental variability. These plots are the operative unit of replication in our sample-size sensitivity analysis. Only plots with valid observations in both years of a pairwise comparison were included.

Let $\mathbf{Y}$ denote the $n \times S$ matrix of continuous abundances, where $n$ is the number of plot–year observations and $S$ the number of species present in the site’s sample frame. We fit the **Generalized Joint Attribute Model** (GJAM; Clark et al., 2017) for each site, specifying a continuous abundance (CA) likelihood for all species. The model included fixed effects for **year** (treated as a categorical factor to enable direct pairwise contrasts between years) and **nlcdClass** (absorbing coarse habitat differences). In model notation,

$$
\mathbf{Y} \sim \text{year} + \text{nlcdClass},
$$

with both predictors applied to all species in the multivariate response. We set `REDUCT = FALSE` to retain the full residual covariance structure, allowing estimation of interspecific correlations without dimension reduction. Models were fit using the `fit_gjam_model_test` implementation, which prepares NEON data for GJAM, ensures predictor alignment across species, and applies internal checks for missing or zero-variance responses. MCMC settings followed defaults used in Clark’s published CA examples, providing stable posterior estimates given the sample sizes in this analysis.

## Posterior prediction and detection probability

Posterior predictions were obtained with a custom `manual_posterior_predict()` function, which draws from the joint posterior distribution of regression coefficients and residual covariances produced by a single fitted model. This approach allows new covariate matrices $\mathbf{X}^\ast$ to be evaluated **without refitting** the model, preserving (1) the species–species covariances estimated in the original fit and (2) the full model uncertainty from that fit. By avoiding refitting, we ensure that any differences between baseline and changed scenarios are attributable solely to the imposed change in covariates, not to variation introduced by separate MCMC runs.

For each site, we identified all consecutive year pairs in the record. For each pair, two prediction scenarios were evaluated:

1. **Baseline:** Observed covariates for all plots with valid data in both years.
2. **Changed:** The same plots and covariates, except that for a single focal species, the abundance was increased by 20% relative to baseline. This was implemented by multiplying the predicted baseline abundance for that species by 1.2 **within the design matrix supplied to posterior prediction**, rather than altering the raw training data.

For each species $s$, the **detection probability** $D_s$ was defined as the posterior probability that predicted abundance under the changed scenario exceeded that under baseline by at least the 20% threshold. Detection probabilities were summarized as posterior means and 95% credible intervals.

## Sample-size sensitivity

We repeated the detection probability analysis under reduced sample sizes to assess the effect of plot number on change detection. NEON’s sampling design fixes the maximum number of Distributed Base plots per site in each year according to protocol (Barnett et al., 2019b), but this number can be reduced in practice by weather, access constraints, or operational limits. For a target sample size $N$, the number of plots used for each year pair was

$$
n = \min(N, n_{\text{available}}),
$$

where $n_{\text{available}}$ is the number of Distributed Base plots meeting protocol requirements for both years. Reduced-plot scenarios were generated by random subsampling without replacement from the available plot set. The **baseline** for sample-size sensitivity used $n = n_{\text{available}}$ and represents the maximum achievable detection with the existing data and sampling effort. This framing links the simulation directly to NEON’s field reality, where plot-level replication is the operative sampling unit for detecting change.

## Community-weighted mean detection

To summarize detection at the community level, we calculated the **community-weighted mean** (CWM) detection probability for each year pair and sample size. Relative cover was used only in this step: let $w_s$ be the mean relative cover of species $s$ across the two years (averaged across sampled plots), and $D_s$ the species-level detection probability. The CWM is given by

$$
\text{CWM} = \frac{\sum_s w_s D_s}{\sum_s w_s}.
$$

Credible intervals for the CWM were obtained by repeating this calculation for each posterior draw, preserving the joint uncertainty across species.

## Decision criterion

For each sample size $N$, we report the fraction of year pairs whose CWM detection probability met or exceeded 0.8. This threshold provides an operational criterion for adequate monitoring performance. Species-specific results are retained for site-level examples in the Results section to illustrate variation in detectability across functional groups and abundances.


# Results

We summarize results in four main steps, each tied directly to the methods:  
(1) species-level detection probabilities,  
(2) posterior distributions for an example species × year pair,  
(3) community-weighted detection for the example site, and  
(4) site-level decision summaries that connect sample-size scenarios to baseline performance.  
This progression mirrors the modeling workflow and culminates in a final table (Table 1) linking detection capacity to sample-size recommendations.

## Species-level detection (example site)

We begin at the species level, evaluating how detection probability changes with sample size for the most abundant taxa at a representative site.  
For each year pair, we plot mean detection probabilities with 95% credible intervals, derived from posterior draws of the fitted GJAM model without re-fitting under simulated changes.  
Dashed horizontal lines indicate the target threshold (0.8), and dotted overlays show baseline detection for the same species–year pairs.  
This allows direct visual comparison between the current sampling design and reduced-plot scenarios.  
Species are ranked by their relative contribution to total community cover, ensuring that the species shown are those most influential to community change metrics.

```{r species-top, warning=FALSE, message=FALSE, fig.cap='Figure 1. Species-level detection (variable mode) with baseline overlays where available; example site shows top species by relative cover. Points/lines are replicate means; ribbons, 95% credible intervals.'}


# pick top species at example site by relative cover
top_species <- relative_cover_df %>%
  rename(site = siteID, species = taxonID) %>%
  filter(site == example_site) %>%
  group_by(species) %>%
  summarise(mean_rc = mean(relative_cover, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(mean_rc)) %>%
  slice_head(n = params$top_n_species) %>%
  pull(species)

# species-level summary: mean + CI (variable view)
# full_detection_summary already species-level; normalize names if needed
fds <- full_detection_summary %>%
  mutate(
    year_pair = dplyr::coalesce(.data$year_pair, paste0(.data$year_baseline, "_", .data$year_changed))
  )
if ("detect_prob" %in% names(fds) && !"mean_detection" %in% names(fds)) {
  fds <- fds %>% rename(mean_detection = detect_prob)
}
if ("requested_sample_size" %in% names(fds)) {
  fds <- fds %>% mutate(sample_size = requested_sample_size)
}

var_sum <- fds %>%
  filter(site == example_site, species %in% top_species, sample_size >= MIN_N) %>%
  group_by(species, year_pair, sample_size) %>%
  summarise(
    mean_det = mean(mean_detection, na.rm = TRUE),
    lo       = mean(ci_lower, na.rm = TRUE),
    hi       = mean(ci_upper, na.rm = TRUE),
    .groups  = "drop"
  )

p <- ggplot(var_sum, aes(sample_size, mean_det, color = species, group = species)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = lo, ymax = hi), width = 0.3, alpha = 0.5) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(breaks = sort(unique(var_sum$sample_size))) +
  facet_wrap(~ year_pair, scales = "free_x") +
  labs(x = "Sample size (plots)", y = "Detection probability") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

# Baseline overlays (species-level means) if available
if (!is.null(cdb)) {
  base_pts <- cdb %>%
    filter(site == example_site, species %in% top_species, sample_size >= MIN_N) %>%
    transmute(species, year_pair, sample_size, y = mean_detection)
  if (nrow(base_pts)) {
    p <- p + geom_point(data = base_pts, aes(x = sample_size, y = y),
                        inherit.aes = FALSE, shape = 21, size = 2.2, stroke = 0.8) +
      geom_line(data = base_pts, aes(x = sample_size, y = y, group = species),
                inherit.aes = FALSE, linetype = "dotted", alpha = 0.6)
  }
}
p

```

## Posterior distribution example (per-draw CWM)

To illustrate how detection probability is propagated from species-level results to the community scale, Figure 2 shows the posterior difference distribution for one example species × year pair at the example site.
The histogram summarizes community-weighted detection across posterior draws, with each draw weighted by the species’ average relative cover between the two years.
The vertical dashed line marks a simulated +20% increase in cover.
This visualization highlights the uncertainty retained from the original GJAM fit — avoiding model re-fitting preserves the posterior covariance structure among species and maintains computational efficiency.

```{r posterior-example, warning=FALSE, message=FALSE, fig.cap='Figure 2. Posterior difference distribution for an example species × year pair at the example site. Dashed line marks a +20% increase.'}
# Helper to choose an example species × pair present in summaries

if (!has_draws_idx_var) {
  plot.new(); title("No draws index available."); 
} else {
  # choose an index row with N >= MIN_N for the example site (filenames may not encode site; join via year_pair/replicate)
  idx <- draws_idx_variable %>%
    left_join(cd_var %>% distinct(site, year_pair, replicate, sample_size),
              by = c("year_pair","replicate")) %>%
    filter(site == example_site, sample_size >= MIN_N) %>%
    arrange(sample_size)
  if (!nrow(idx)) { plot.new(); title("No qualifying draws for example site and N ≥ min."); } else {
    fp <- idx$file[[1]]
    dd <- tryCatch(arrow::read_parquet(fp), error = function(e) NULL)
    if (is.null(dd) || !"detected" %in% names(dd)) {
      plot.new(); title("Draws file found, but missing `detected` column.")
    } else {
      # average relative cover across plots per species & year
      rc <- relative_cover_df %>%
        rename(site = siteID, species = taxonID) %>%
        group_by(site, species, year) %>%
        summarise(mean_relative_cover = mean(relative_cover, na.rm = TRUE), .groups = "drop")

      # compute draw-level CWM distribution
      dist_df <- dd %>%
        mutate(
          year_baseline = as.integer(year_baseline),
          year_changed  = as.integer(year_changed)
        ) %>%
        left_join(rc, by = c("site","species","year_baseline"="year")) %>% rename(rc_b = mean_relative_cover) %>%
        left_join(rc, by = c("site","species","year_changed" ="year")) %>% rename(rc_c = mean_relative_cover) %>%
        mutate(w = rowMeans(cbind(rc_b, rc_c), na.rm = TRUE)) %>%
        filter(!is.na(w)) %>%
        mutate(weighted = as.numeric(detected) * w) %>%
        group_by(draw) %>%
        summarise(cwm_draw = sum(weighted, na.rm = TRUE) / sum(w, na.rm = TRUE), .groups = "drop")

      ggplot(dist_df, aes(x = cwm_draw)) +
        geom_histogram(bins = 40) +
        geom_vline(xintercept = threshold, linetype = "dashed") +
        scale_x_continuous(limits = c(0,1)) +
        labs(x = "Community-weighted detection (per draw)", y = "Count")
    }
  }
}

```

## Community-weighted detection by year pair (example site)

Moving from species-level to community-level detection, Figure 3 presents community-weighted mean (CWM) detection probabilities by year pair for the example site.
For each year pair, detection probability is averaged across species using relative-cover weights.
The dashed line indicates the target detection threshold (0.8), enabling quick assessment of whether a given sample size meets the desired detection probability for that year pair.
Baseline overlays allow direct comparison between current and reduced sampling.

```{r cwm-by-pair, warning=FALSE, message=FALSE, fig.cap='Figure 3. Community-weighted detection by year pair (variable mode) at the example site.'}
p <- ggplot(rep_avg_var_site, aes(x = sample_size, y = mean_detection, color = year_pair, group = year_pair)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(breaks = sort(unique(rep_avg_var_site$sample_size))) +
  labs(x = "Sample size (plots)", y = "Detection probability", color = "Year pair") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
add_baseline_layer(p, rep_avg_base_site, yvar = "mean_detection")

```

## Mean across year pairs (example site)

While Figure 3 disaggregates results by year pair, Figure 4 condenses these into a single mean detection probability per sample size, averaged across all year pairs at the example site.
This site-level average provides a straightforward summary of how detection capacity scales with plot number, relative to the baseline design.

```{r cwm-mean, warning=FALSE, message=FALSE, fig.cap='Figure 4. Mean community-weighted detection across year pairs (variable mode) at the example site.'}
p <- ggplot(site_avg_var_site, aes(x = sample_size, y = detection_mean, group = 1)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(breaks = sort(unique(site_avg_var_site$sample_size))) +
  labs(x = "Sample size (plots)", y = "Detection probability") +
  theme_minimal(base_size = 12)
add_baseline_layer(p, site_avg_base_site, yvar = "detection_mean")

```

## Site-level decision summary (variable mode)

Finally, Figure 5 and Table 1 synthesize the variable-sample-size results across all sites.
Figure 5 shows, for each site, the fraction of year pairs that meet or exceed the target detection threshold (0.8) for a given sample size.
Table 1 lists these fractions alongside the mean baseline detection at the same N, allowing direct evaluation of whether reduced sampling could maintain acceptable detection performance (≥ 75% of year pairs meeting the threshold).

```{r decision-summary, warning=FALSE, message=FALSE, fig.cap='Figure 5. Fraction of year pairs meeting the 0.8 detection threshold by requested sample size (variable mode).'}
frac_tbl <- fraction_passing(rep_avg_var, threshold = threshold) %>%
  arrange(site, sample_size)

ggplot(frac_tbl, aes(x = sample_size, y = prop_yearpairs_passing, group = site)) +
  geom_line(alpha = 0.4) +
  geom_point(alpha = 0.6, size = 1.8) +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(breaks = sort(unique(frac_tbl$sample_size))) +
  labs(x = "Sample size (plots)", y = "Fraction of year pairs ≥ 0.8") +
  theme_minimal(base_size = 12)

# Table: add baseline comparison when available (matching by site × N)
var_pass <- frac_tbl %>%
  rename(fraction_yearpairs_passing = prop_yearpairs_passing,
         mean_detection_at_N        = mean_at_size,
         sd_detection_at_N          = sd_at_size)

base_compact <- if (!is.null(site_avg_base)) site_avg_base %>%
  select(site, sample_size, baseline_detection_at_N = detection_mean) else tibble()

site_table <- var_pass %>%
  left_join(base_compact, by = c("site","sample_size")) %>%
  arrange(site, sample_size)

knitr::kable(
  site_table,
  caption = "Table 1. Fraction of year pairs meeting the 0.8 detection threshold by requested sample size (variable mode), with baseline mean detection at the same N when available."
)


```

# Discussion

Our results quantify how NEON’s distributed plant diversity plots perform in detecting interannual changes in species cover and in aggregated community-weighted metrics. Baseline overlays reveal the “headroom” of the current design—the detection probability achieved with the actual number of plots sampled—while variable-$N$ curves show how detection probability changes as sampling effort is reduced. This side-by-side comparison makes it clear where reduced sampling could still meet monitoring goals and where the current effort is essential.

At the species level, responses to a simulated 20% change vary widely. This heterogeneity reflects differences in abundance, spatial distribution, and covariance with other species. Weighting by relative cover to produce community-weighted means integrates these heterogeneous species signals into a clearer, site-level metric, while still preserving the propagation of model uncertainty from the posterior.

Finally, the decision-focused framing—expressed as the fraction of year pairs meeting a pre-specified detection criterion (here, 0.8)—links statistical results directly to operational goals. By using the same model fit for both baseline and reduced-$N$ scenarios, we maintain the covariance structure among species and the uncertainty captured in the original model, avoiding the noise and bias that can result from refitting models to perturbed datasets.


# Key Findings

1. **Current design performance varies across sites.** Some sites already exceed the 0.8 detection threshold for most year pairs, while others fall short even at baseline.
2. **Species-level detection is heterogeneous.** Many abundant species achieve high detection probabilities, but rarer or patchier species have lower and more variable detection rates.
3. **Community-weighted means clarify the big picture.** Weighting by relative cover yields a stable, interpretable site-level metric that integrates species-specific detection probabilities and uncertainty.
4. **Reduced sampling can be efficient—but only at some sites.** Sensitivity curves reveal that certain sites can meet or approach baseline detection with fewer plots, while others require the full current design to achieve target performance.
5. **Method preserves spatial covariance and posterior uncertainty.** By simulating change without altering raw data or refitting, the analysis maintains the integrity of the fitted joint species model and its inferred species-to-species relationships.


# Conclusion

This analysis demonstrates how a joint species modeling framework can translate NEON’s distributed plant diversity data into site-specific, decision-ready metrics of sampling performance. By quantifying detection probabilities for both individual species and community-weighted signals, we connect raw observations to operational thresholds for monitoring change. Our results highlight that the current distributed plot design often achieves high interannual detection power, but with uneven performance across sites and species. Sensitivity curves provide a transparent way to evaluate the trade-offs of reduced sampling effort, revealing where efficiency gains may be possible and where current effort is necessary to meet monitoring goals. Because the framework propagates posterior uncertainty and maintains the full covariance structure from a single fitted model, it provides a robust basis for guiding long-term design decisions, adaptive management, and scenario testing across NEON and similar large-scale ecological networks.


# Appendix {-}

## A.1 All sites: CWM detection by year pair

```{r app-all-by-yearpair, fig.cap='Figure A1. Community-weighted detection by year pair for all sites (variable mode).'}
ggplot(rep_avg_var, aes(x = sample_size, y = mean_detection, color = year_pair, group = year_pair)) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(breaks = sort(unique(rep_avg_var$sample_size))) +
  labs(x = "Sample size (plots)", y = "Detection probability", color = "Year pair") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")

```

## A.2 All sites: Mean across year pairs

```{r app-all-mean, fig.cap='Figure A2. Mean community-weighted detection across year pairs for all sites (variable mode).'}
ggplot(site_avg_var, aes(x = sample_size, y = detection_mean, group = site)) +
  geom_line(alpha = 0.6) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = threshold, linetype = "dashed") +
  scale_y_continuous(limits = c(0,1)) +
  scale_x_continuous(breaks = sort(unique(site_avg_var$sample_size))) +
  labs(x = "Sample size (plots)", y = "Detection probability") +
  theme_minimal(base_size = 12)

```


##References
- Barnett, D. T., et al. (2019a). The plant diversity sampling design for the National Ecological Observatory Network. *Ecosphere*, 10(2), e02603.
- Barnett, D. T., et al. (2019b). The terrestrial organism and biogeochemistry spatial sampling design for the National Ecological Observatory Network. *Ecosphere*, 10(2), e02604.
- Clark, J. S., et al. (2017). Generalized joint attribute modeling for biodiversity analysis: Median-zero, multivariate, multifarious data. *Ecological Monographs*, 87(1), 34–56.

